---
title: DE on RNA-seq data using kallisto and sleuth to obtain mean expression to use
  as weight for the Lancaster method
output:
  html_document:
    df_print: paged
---

## Load libraries
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(dplyr)
library(readr)
library(biomaRt)
library(data.table)
library(aggregation)
library(remotes)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

source R profile. Memory was set to 500000.
```{r, include=FALSE, echo=FALSE}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Set wd
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```

## Load data
```{r}
transcripts <- read_csv("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_RNAseq_transcripts.csv")

#METADATA
sample_info <- read_csv("/Users/castilln/Desktop/thesis/localdata/depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("/Users/castilln/Desktop/thesis/localdata/depmap/mutations_spliceosome.csv")
```

Load lm results from transcripts_lm
```{r}
lm_res = readRDS("../github/results/differential_expression/transcripts/lm_res_symbols.rds")

head(lm_res)
```

Contingency table - Annotation of cell lines with spliceosome mutations
```{r}
cell_lines_list = 
  CCLE_mutations %>% 
  select(stripped_cell_line_name, primary_disease, DepMap_ID) 

cell_lines_list_mutated = 
 cell_lines_list %>% 
   mutate(spliceosome_mutated = 
           case_when(
            cell_lines_list$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !cell_lines_list$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) %>% 
  distinct()

cell_lines_list_mutated %>% 
  distinct() %>%
  group_by(primary_disease,spliceosome_mutated) %>%
  tally()

table(distinct(cell_lines_list_mutated)$primary_disease,distinct(cell_lines_list_mutated)$spliceosome_mutated)

spliceosome_mutated_ID = 
  cell_lines_list_mutated %>% 
  select(DepMap_ID, spliceosome_mutated) %>% 
  distinct() 
```

```{r}
#RENAME FIRST VARIABLE 
transcripts = 
  transcripts %>% 
  rename("DepMap_ID" = "X1")

#JOIN TRANSCRIPT INFO WITH METADATA
long_transcripts_info = 
  transcripts %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "Gene", values_to = "log_TPM") %>% 
  left_join(cell_lines_list_mutated, by = "DepMap_ID")

head(long_transcripts_info)
```

## Pre-processing 



Create a new column for ensembl transcript ID 
```{r}
median_counts = 
  median_counts %>%
    separate(col = "Gene", into = c("Hugo_Symbol", "ensembl_id"), sep = "\\s") 
   
median_counts$ensembl_id <- gsub("\\(|\\)","",as.character(median_counts$ensembl_id))

counts_lm = 
  median_counts %>% 
  select(Hugo_Symbol, ensembl_id, stripped_cell_line_name, primary_disease, spliceosome_mutated, log_TPM)

##CHECK THAT THERE ARE NO DUPLICATES
#counts_lm %>% 
#distinct() %>% 
#dim()
```

Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
keep <- c("Bone Cancer", "Fibroblast", "Leukemia", "Lung Cancer", "Ovarian Cancer", "Pancreatic Cancer", "Sarcoma", "Skin Cancer")

df_deseq = 
counts_lm %>% 
  filter(primary_disease %in% keep)

head(df_deseq)
```

Filter out those genes whose median for expression is lower than 1 across all samples
```{r}
median_counts =
  df_deseq %>% 
  group_by(Gene) %>% 
  dplyr::mutate(median = median(log_TPM, na.rm=TRUE)) %>% 
  filter(median > 1)
```

For lancaster method, we need to use the mean expression for each transcript in order to weight each p-value
```{r}
mean_spliceosome = 
  df_deseq %>% 
  group_by(spliceosome_mutated, ensembl_id) %>% 
  summarise_at(vars(log_TPM), funs(mean(., na.rm = TRUE)))

mean_disease = 
  df_deseq %>% 
  group_by(primary_disease, ensembl_id) %>% 
  summarise_at(vars(log_TPM), funs(mean(., na.rm = TRUE)))
```

```{r}
mean_spliceosome_yes = 
  mean_spliceosome %>%
  filter(spliceosome_mutated =="YES") %>% 
  rename("weight" = "log_TPM") %>% 
  ungroup() %>% 
  select(-spliceosome_mutated)
```

```{r}
head(mean_disease)
```

Filter spliceosome mutated from linear model results 
```{r}
lm_spl = 
  lm_res %>% 
  filter(term == "spliceosome_mutatedYES")
  
head(lm_spl)
```

Join weights to correspondent p-value 
```{r}
lanc = 
  lm_spl %>% 
  left_join(mean_spliceosome_yes, by = "ensembl_id") %>% 
  select(Hugo_Symbol, ensembl_id, p.value, weight) 

head(lanc)
```

## Lancaster method

Let's do lancaster aggregation looping thorugh the list of p-values obtained from the linear model
```{r}
##CREATE A LIST WITH ALL GENES WITH MORE THAN 1 ISOFORM
list_symbols = 
  lm_spl %>% 
  count(Hugo_Symbol) %>% 
  rename("n_transcripts" = "n") %>% 
  filter(n_transcripts > 1) %>% 
  select(Hugo_Symbol) %>% 
  as.data.frame()
  

##CREATE A DF TO STORE AGGREGATED P VALUES 
n <- lengths(list_symbols)

  #PREDEFINED DIMENSIONS
df_ag = data.table(Hugo_Symbol = rep(as.character(NA), n),
                   p.value = rep(0,n))

##LOOP THROUGH HUGO SYMBOLS AND AGGREGATE THE P VALUES
for (i in 1:n) {
  ##GET HUGO_SYMBOL
  temp_symbol = list_symbols[i,1]

  ##EXTRACT P-VALUES AND WEIGHT FOR CORRESPONDING HUGO SYMBOL (IN LOOP)
  temp = 
    lanc %>% 
    filter(Hugo_Symbol == temp_symbol) %>% 
    select(c("p.value", "weight"))
  
  ##CREATE VECTOR OF P VALUES
  temp_p = 
    temp %>% 
    pull(p.value)
  
  ##CREATE VECTOR OF WEIGHTS
  temp_w = 
    temp %>% 
    pull(weight)
  
  ##LANCASTER
  temp_agg = 
    lancaster(temp_p, temp_w)
  
  ##STORE IN DATA TABLE
  df_ag[i, Hugo_Symbol := as.character(temp_symbol)]
  df_ag[i, p.value := temp_agg]
}

head(df_ag)

```

Get p-values of those genes w/o isoforms
```{r}
p.val_no_iso = 
  lanc %>% 
  select(-weight) %>% 
  anti_join(df_ag, by = "Hugo_Symbol") %>% 
  select(-ensembl_id)
```

Join with aggregated p-values 
```{r}
##DF CONTAINING ALL THE P-VALUES
all_p.val = 
  rbind(df_ag, p.val_no_iso)
```

Save p.values as RSD
```{r}
saveRDS(all_p.val, file = "/Users/castilln/Desktop/thesis/github/results/linear_models/transcripts/p_values_lancaster.rds")
```

