---
title: "PIANO analysis"
output: html_notebook
---
# Load libraries
```{r}
library(tidyverse)
library(piano)
library(edgeR)
```

Set wd
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```


Load lm results from transcripts_lm
```{r}
lm_res = readRDS("../github/results/linear_models/transcripts/lm_res_symbols.rds")

head(lm_res)
```

Load aggregated p-values
```{r}
lanc_p_values = readRDS("../github/results/linear_models/transcripts/p_values_lancaster.rds")

head(lanc_p_values)
```
Load gene sets 
```{r}
kegg_gsc = loadGSC(file="gene_sets/c2.cp.kegg.v7.2.symbols.gmt")
hallmarks_gsc = loadGSC(file="gene_sets/h.all.v7.2.symbols.gmt")
```


# Comp. YesvsNo

FDR correction
```{r}
lm_res %>% 
  filter(term == "spliceosome_mutatedYES") %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(p.value) %>% 
  head()
```

```{r}
lm_adj = 
  lm_res %>% 
  filter(term == "spliceosome_mutatedYES") %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(p.value) 

head(lm_adj)
```

Extract FDR adjusted p-values and logFC (estimate)
```{r}
p_val = 
  lm_adj %>% 
  pull(p.value)

p_adj = 
  lm_adj %>% 
  pull(fdr)

log2fc = 
  lm_adj %>% 
  pull(estimate)


names(p_val) = 
  lm_adj %>% 
  pull(Hugo_Symbol)

names(p_adj) = 
  lm_adj %>% 
  pull(Hugo_Symbol)

names(log2fc) =
  lm_adj %>% 
  pull(Hugo_Symbol)
```

```{r}
cat("p-values \n")
head(p_adj)
```

```{r}
cat("\nlog2-FC \n")
head(log2fc)
```

Run PIANO
For kegg gs.
```{r}
library(snowfall)
gsa_adj_kegg <- runGSA(p_adj, 
                  log2fc,
                  gsc = kegg_gsc,
                  ncpus=8, 
                  geneSetStat = "reporter",
                  signifMethod = "nullDist",
                  nPerm = 10000)

gsa_adj_kegg
```

```{r, fig.width=8, fig.height=8}
GSAheatmap(gsaRes = gsa_adj_kegg)
```

Run PIANO on KEGG gs, using non-adj p-values 
```{r}
gsa_kegg <- runGSA(p_val, 
                  log2fc,
                  gsc = kegg_gsc,
                  ncpus=8, 
                  geneSetStat = "reporter",
                  signifMethod = "nullDist",
                  nPerm = 10000)

```

```{r, fig.width=8, fig.height=8}
GSAheatmap(gsaRes = gsa_kegg)
```
Run PIANO on KEGG gs, using gene sampling to assess significance 
```{r}
gsa_kegg_geneS <- runGSA(p_val, 
                  log2fc,
                  gsc = kegg_gsc,
                  ncpus=8, 
                  geneSetStat = "reporter",
                  signifMethod = "geneSampling",
                  nPerm = 10000)

```

```{r, fig.width=8, fig.height=8}
GSAheatmap(gsaRes = gsa_kegg_geneS)
```

Run PIANO on hallmarks gsc, theoretical null distribution.
```{r}
gsa_res <- runGSA(p_adj, 
                  log2fc,
                  gsc = hallmarks_gsc,
                  ncpus=8, 
                  geneSetStat = "reporter",
                  signifMethod = "nullDist",
                  nPerm = 10000)
```

```{r, fig.width=8, fig.height=8}
GSAheatmap(gsaRes = gsa_res)
```
Run PIANO on hallmarks gsc, gene sampling.
```{r}
gsa_res <- runGSA(p_adj, 
                  log2fc,
                  gsc = hallmarks_gsc,
                  ncpus=8, 
                  geneSetStat = "reporter",
                  signifMethod = "geneSampling",
                  nPerm = 10000)
```

```{r, fig.width=8, fig.height=8}
GSAheatmap(gsaRes = gsa_res)
```


# saveRDS
```{r}

```



