---
title: "Drug  sensitivity"
output: html_notebook
---

Load libraries
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(readr)
library(biomaRt)
library(data.table)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Set wd
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```

Load data
```{r}
drug_df <- fread("depmap/primary-screen-replicate-collapsed-logfold-change.csv", header = TRUE) 

#METADATA
sample_info <- read_csv("depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("depmap/mutations_spliceosome.csv")
```

Rename V1 to DepMap_ID
```{r}
drug_df = 
  drug_df %>% 
  rename("DepMap_ID" = "V1")
```

Create meta data of interest
```{r}
meta = 
  sample_info %>% 
  select(c("DepMap_ID", "stripped_cell_line_name", "primary_disease"))
```

Join tidy crispr data and join with metadata
```{r}
drug_meta = 
  drug_df %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "drug", values_to = "value") %>% 
  left_join(meta, by = "DepMap_ID")
```

Annotate mutations in the spliceosome
```{r}
drug_meta = 
 drug_meta %>% 
   mutate(spliceosome_mutated = 
           case_when(
            drug_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !drug_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) 
```


```{r}
table(distinct(drug_meta)$primary_disease,distinct(drug_meta)$spliceosome_mutated)
```

## Pre-processing: quality check. 
Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
keep <- c("Bone Cancer", "Bile Duct Cancer", "Brain Cancer", "Head and Neck Cancer",  "Lung Cancer", "Pancreatic Cancer")

df_filt = 
  drug_meta %>% 
  filter(primary_disease %in% keep) %>% 
  drop_na()


##CHECK THERE ARE NO DUPLICATES
#df_filt %>% 
 # dim()

#df_filt %>% 
# distinct() %>% 
# dim()
```

```{r}
head(df_filt)
```


QQ plots
```{r}
theme_set(
  theme_minimal() +
    theme(legend.position = "top")
  )

##PLOT 
ggplot(df_filt, aes(sample = value)) + 
  facet_wrap(facets = "primary_disease") +
  stat_qq() + 
  stat_qq_line() 
```
```{r}
head(df_filt)
```


Lets try to standardize the data 
```{r}
st_df_filt = 
  df_filt
```


## Wilcoxon-test
Estimate significance using a non-parametric method
Prepare data for Wilcoxon 
```{r, message=FALSE}
df_filt %>% 
  group_by(spliceosome_mutated) %>% 
  summarise(
    count = n(),
    median = median(value, na.rm = TRUE),
    IQR = IQR(value, na.rm = T)
    )
```

Visualize data 
```{r}
ggboxplot(df_filt, x = "spliceosome_mutated", y = "value", 
          color = "spliceosome_mutated", palette = c("mediumvioletred","midnightblue"), 
          ylab = "Gene effect", xlab = "Defective spliceosome")
```

2 samples Wilcoxon test
```{r}
res <- wilcox.test(value  ~ spliceosome_mutated, data = df_filt, exact = FALSE)

res
```
Print p-value 
```{r}
res$p.value
```

# Wilcoxon test per cell line

Fit directly to linear model
```{r}
library(broom)
wilc_res =
  df_filt %>% 
  ##NEST BY CELL LINE
  nest(-drug) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied) %>% 
  select(-c("data", "model"))


head(wilc_res)
```

FDR 
```{r}
wilc_res %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(p.value) %>% 
  head()
```
Repeat wilcoxon test per cancer - FIX ERROR. NOT ALL DRUGS HAVE BOTH FACTORS
```{r}
##LIST OF CANCERS
list_cancer = 
  df_filt %>% 
  pull(primary_disease) %>% 
  unique()

##EMPTY LIST TO STORE LINEAR MODELS
wilc_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  wilc_list[[paste0("wilc.", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-drug) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```

```{r}
print(list_cancer)

```

Pancreatic cancer
```{r}
wilc_list$`lm.Pancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```


t-test
```{r}
library(broom)
t_res =
  df_filt %>% 
  ##NEST BY CELL LINE
  nest(-drug) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~t.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied) %>% 
  select(-c("data", "model"))


head(t_res)
```


Repeat t-test per cancer
```{r}
##EMPTY LIST TO STORE LINEAR MODELS
t_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  t_list[[paste0("t", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-Hugo_Symbol) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~t.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```
