---
title: "Kolmogorov-Smirnov test in data from PRISM screening"
output: html_notebook
---

Description: test statistical significant differences on drug repsonses between spliceosome-wt cell lines and spliceosome-mutated cell lines using K-S test. 

Libraries 
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(readr)
library(data.table)
library(dgof)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Set wd
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```

Load data
```{r}
drug_df <- fread("depmap/drug_sensitivity/primary-screen-replicate-collapsed-logfold-change.csv", header = TRUE) 

#METADATA
sample_info <- read_csv("depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("depmap/mutations_spliceosome.csv")
```

Rename V1 to DepMap_ID
```{r}
drug_df = 
  drug_df %>% 
  rename("DepMap_ID" = "V1")
```

Create meta data of interest
```{r}
meta = 
  sample_info %>% 
  select(c("DepMap_ID", "stripped_cell_line_name", "primary_disease"))
```

Join tidy crispr data and join with metadata
```{r}
drug_meta = 
  drug_df %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "drug", values_to = "value") %>% 
  left_join(meta, by = "DepMap_ID")
```

Annotate mutations in the spliceosome
```{r}
drug_meta = 
 drug_meta %>% 
   mutate(spliceosome_mutated = 
           case_when(
            drug_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !drug_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) 
```


```{r}
table(distinct(drug_meta)$primary_disease,distinct(drug_meta)$spliceosome_mutated)
```

## Pre-processing: quality check. 
Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
keep <- c("Bone Cancer", "Bile Duct Cancer", "Brain Cancer", "Head and Neck Cancer",  "Lung Cancer", "Pancreatic Cancer")

df_filt = 
  drug_meta %>% 
  filter(primary_disease %in% keep) %>% 
  drop_na()


##CHECK THERE ARE NO DUPLICATES
#df_filt %>% 
 # dim()

#df_filt %>% 
# distinct() %>% 
# dim()
```

```{r}
head(df_filt)
```

Transform primary disease as factor
```{r}
df_filt = 
  df_filt %>% 
  mutate(primary_disease = as.factor(primary_disease))
```

Kolmogorov-Smirnov test
```{r}
#SEPARATE THE DATAFRAME INTO TWO GROUPS
df_no = df_filt[df_filt$spliceosome_mutated == "NO",]
df_yes = df_filt[df_filt$spliceosome_mutated == "YES",]

#WT SPLICEOSOME
#TRASNFORM TO WIDE FORMAT (SYMBOLS AS COLUMNS)
df_no_wide = 
  df_no %>% 
  pivot_wider(id_cols = DepMap_ID, names_from = drug) %>% 
  as.data.frame()

#DROP NA VALUES
df_no_wide <- drop_na(df_no_wide)

#SET DEPMAP ID AS ROWNAMES
rownames(df_no_wide) <- df_no_wide$DepMap_ID
df_no_wide <- df_no_wide[,-1]

#MUTATED SPLICEOSOME
#TRASNFORM TO WIDE FORMAT (SYMBOLS AS COLUMNS)
df_yes_wide = 
  df_yes %>% 
  pivot_wider(id_cols = DepMap_ID, names_from = drug) %>% 
  as.data.frame()

#DROP NA VALUES
df_yes_wide <- drop_na(df_yes_wide)

#SET DEPMAP ID AS ROWNAMES
rownames(df_yes_wide) <- df_yes_wide$DepMap_ID
df_yes_wide <- df_yes_wide[,-1]


df_ks <- data.frame(Symbol = character(4686), D = numeric(4686), p = numeric(4686), stringsAsFactors = F)
```

Loop for KS test --- pan-cancer
```{r}
for (j in 1:4686){  
  k <- ks.test(df_no_wide[,j], df_yes_wide[,j])  
  df_ks$Symbol[j] <- names(df_no_wide)[j]
  df_ks$D[j] <- k$statistic
  df_ks$p[j] <- k$p.value
}
```

FDR correction
```{r}
df_ks %>% 
  mutate(fdr = p.adjust(p, method = "fdr")) %>% 
  filter(p < 0.05)
```

