---
title: "Drug  sensitivity - TCGA core splicing factors"
output: html_notebook
---

Load libraries
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(readr)
library(biomaRt)
library(data.table)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Set wd
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```

Load data
```{r}
drug_df <- fread("depmap/drug_sensitivity/primary-screen-replicate-collapsed-logfold-change.csv", header = TRUE) 

#METADATA
sample_info <- read_csv("depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("spliceosome/tcga_core_spliceosome_ccle_mutations.csv")

head(mutations_spliceosome)
```

Rename V1 to DepMap_ID
```{r}
drug_df = 
  drug_df %>% 
  rename("DepMap_ID" = "V1")
```

Create meta data of interest
```{r}
meta = 
  sample_info %>% 
  select(c("DepMap_ID", "stripped_cell_line_name", "primary_disease"))
```

Join tidy crispr data and join with metadata
```{r}
drug_meta = 
  drug_df %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "drug", values_to = "value") %>% 
  left_join(meta, by = "DepMap_ID")
```

Annotate mutations in the spliceosome
```{r}
drug_meta = 
 drug_meta %>% 
   mutate(spliceosome_mutated = 
           case_when(
            drug_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !drug_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) 
```


```{r}
table(distinct(drug_meta)$primary_disease,distinct(drug_meta)$spliceosome_mutated)
```

## Pre-processing: quality check. 
Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
no_keep <- c("Bile Duct Cancer", "Endometrial/Uterine Cancer", "Esophageal Cancer", "Fibroblast", "Gallbladder Cancer", "Prostate Cancer")

'%notin%' <- function(x,y)!('%in%'(x,y))

df_filt = 
  drug_meta %>% 
  filter(primary_disease %notin% no_keep) %>% 
  drop_na()


##CHECK THERE ARE NO DUPLICATES
#df_filt %>% 
 # dim()

#df_filt %>% 
# distinct() %>% 
# dim()
```

```{r}
head(df_filt)
```


QQ plots
```{r}
theme_set(
  theme_minimal() +
    theme(legend.position = "top")
  )

##PLOT 
ggplot(df_filt, aes(sample = value)) + 
  facet_wrap(facets = "primary_disease") +
  stat_qq() + 
  stat_qq_line() 
```
```{r}
head(df_filt)
```


Lets try to standardize the data 
```{r}
st_df_filt = 
  df_filt
```


## Wilcoxon-test
Estimate significance using a non-parametric method
Prepare data for Wilcoxon 
```{r, message=FALSE}
df_filt %>% 
  group_by(spliceosome_mutated) %>% 
  summarise(
    count = n(),
    median = median(value, na.rm = TRUE),
    IQR = IQR(value, na.rm = T)
    )
```

Visualize data 
```{r}
ggboxplot(df_filt, x = "spliceosome_mutated", y = "value", 
          color = "spliceosome_mutated", palette = c("red","black"), 
          ylab = "Gene effect", xlab = "Defective spliceosome") + 
  stat_compare_means(method = "wilcox.test", label = "p.format")
```

2 samples Wilcoxon test
```{r}
res <- wilcox.test(value  ~ spliceosome_mutated, data = df_filt, exact = FALSE)

res
```
Print p-value 
```{r}
res$p.value
```

# Wilcoxon test per cell line

Fit directly to linear model
```{r}
library(broom)
wilc_res =
  df_filt %>% 
  ##NEST BY CELL LINE
  nest(-drug) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied) %>% 
  select(-c("data", "model"))


head(wilc_res)
```

FDR 
```{r}
wilc_res %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(p.value) %>% 
  head()
```
Repeat wilcoxon test per cancer 
```{r}
##LIST OF CANCERS
list_cancer = 
  df_filt %>% 
  pull(primary_disease) %>% 
  unique()

##EMPTY LIST TO STORE LINEAR MODELS
wilc_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  wilc_list[[paste0("wilc.", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-drug) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```

Prepare for tSNE/PCA
```{r}
head(df_filt)
##PIVOT WIDER
df_wide = 
  df_filt %>% 
  select(DepMap_ID, drug, value, spliceosome_mutated, primary_disease) %>% 
  pivot_wider(names_from = drug, values_from = value) %>%  
  as.data.frame() %>% 
  relocate(DepMap_ID, .after=primary_disease)

drug_pca <- as.matrix(df_wide[3:ncol(df_wide)])
drug <- as.matrix(df_wide[,4:ncol(df_wide)])
disease <- df_wide[, 2] #disease
mutated <- df_wide[, 1]

##GET RID OF NA VALUES AND SUBSTITUTE FOR MEAN EXPRESSION PER CELL LINE
for(i in 1:ncol(drug)){
  drug[is.na(drug[,i]), i] <- mean(drug[,i], na.rm = TRUE)
} 

#Identify colums with 0 variance and remove them from dataset - in order to be able to perform the PCA
drug_var <- drug[ , which(apply(drug, 2, var) != 0)]
```

Run PCA

```{r,  fig.height=4, fig.width=7}
pca1 = prcomp(drug_var, center = TRUE, scale = TRUE, na.action=na.omit(drug_var))

plotData = pca1$x[,1:2]
plotData = cbind(DepMap_ID = expression_pca[,1], plotData)
rownames(plotData) = NULL

head(plotData)

ID = plotData[,1]

to_join = 
  CCLE_mutations %>% 
  select(DepMap_ID) %>% 
  distinct() %>% 
  left_join(cell_lines_list_mutated, by = "DepMap_ID")
  
plotData %>% 
  as_tibble() %>% 
  left_join(to_join, by ="DepMap_ID") %>%
  mutate(PC1 = as.double(PC1),
         PC2 = as.double(PC2))  -> plotData_gg
```

Plot

```{r,  fig.height=6, fig.width=12}

ggplot(plotData_gg, aes(x = PC1,y = PC2, color = spliceosome_mutated)) +
  geom_point() +
  facet_wrap(vars(primary_disease), scales = "free") + 
  stat_ellipse()  -> gg_drug_mutated



  ggsave("/Users/castilln/Desktop/thesis/figures/tcga_core_gg_drug_mutation.png", device = "png", width = 25, height = 15, units = "cm", dpi = "retina")

ggplot(plotData_gg, aes(x = PC1,y = PC2, color = primary_disease)) +
  geom_point() +
  scale_color_manual(values = c("gainsboro", 'forestgreen', 'red2', 'orange',  'cornflowerblue', 
                'magenta', 'darkolivegreen4',  'indianred1',  'tan4', 'darkblue', 
                'mediumorchid1', 'firebrick4',  'yellowgreen', 'lightsalmon', 'tan3',
                "tan1",  'darkgray','wheat4',  '#DDAD4B',  'chartreuse', 
                 'seagreen1', 'moccasin',   'mediumvioletred', 'seagreen', 'cadetblue1',
                "darkolivegreen1" ,"tan2" ,  "tomato3" , "#7CE3D8", "black", "yellow", "violetred", "blue")) -> gg_drug_disease

  ggsave("/Users/castilln/Desktop/thesis/figures/tcga_core_gg_drug_disease.png", device = "png", width = 25, height = 15, units = "cm", dpi = "retina")

grid.arrange(gg_drug_mutated, gg_drug_disease, ncol = 2)

```

Run tSN
```{r, fig.height=5, fig.width=7}
set.seed(6)
tsne_results <- Rtsne(drug, dims = 2, perplexity = 40, theta = 0.5, check_duplicates = FALSE, pca = TRUE)

#plot

plotData =
  tsne_results$Y %>% 
  as_tibble() %>% 
  mutate(DepMap_ID = ID)
 
plotData = 
  plotData %>% 
  left_join(to_join, by ="DepMap_ID") %>% 
  mutate(V1 = as.double(V1),
         V2 = as.double(V2))
 
plotData %>% 
  ggplot(aes(x = V1,y = V2, color = spliceosome_mutated)) +
  geom_point(aes(alpha=0.5), size = 1) +
  xlab(element_blank()) +
  ylab(element_blank()) +
  facet_wrap(vars(primary_disease)) +
  theme_bw() +
  ggtitle("tSNE") +
  theme(legend.position = "bottom",
        aspect.ratio = 1,
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(face = "bold")) +
  scale_color_manual(values = c('forestgreen', 'red2')) ->tsne_expression

ggsave("/Users/castilln/Desktop/thesis/figures/ore_tcga_tsne_drug.png", device = "png", width = 30, height = 25, units = "cm", dpi = "retina")

print(tsne_expression)
```
Pancreatic cancer
```{r}
wilc_list$`lm.Pancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```


t-test
```{r}
library(broom)
t_res =
  df_filt %>% 
  ##NEST BY CELL LINE
  nest(-drug) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~t.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied) %>% 
  select(-c("data", "model"))


head(t_res)
```


Repeat t-test per cancer
```{r}
##EMPTY LIST TO STORE LINEAR MODELS
t_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  t_list[[paste0("t", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-Hugo_Symbol) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~t.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```