---
title: "Exploration of CLL dataset"
author: "Leticia Castillon"
Rstudio version: "1.3.1093"
output: html_notebook
---
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis")
```

load libraries
```{r, message=FALSE, include=FALSE}
library(tidyverse)
library(readr)
library(biomaRt)
library(cowplot)
library(ggpubr)
library(data.table)
library(ggsci)
library(ggrepel)
library(ggExtra)
library(hrbrthemes)
library(wesanderson)
library(EDASeq)
library(dplyr)
library(goseq)
select = dplyr::select
count = dplyr::count
rename = dplyr::rename
```


## Load data

```{r, message=FALSE, include=FALSE}
CCLE_mutations <- read_delim("localdata/depmap/CCLE.csv", ",", escape_double = FALSE, trim_ws = TRUE)
#head(CCLE_mutations)

sample_info <- read_csv("localdata/depmap/sample_info.csv")
#head(sample_info)
```

Merge both datasets (by DepMap_ID)
```{r}
mutations_info = left_join(CCLE_mutations, sample_info, by = c("DepMap_ID"))
```

Export to csv
```{r}
#write.csv(mutations_info,"/Users/castilln/Desktop/thesis/localdata/depmap/mutations_info", row.names = FALSE)
```


## Data exploration
Count number of mutations per cancer
```{r}
mutations_per_cancer = 
  mutations_info %>% 
  count(primary_disease) %>% 
  rename("freq" = "n")

plot_CCLE = 
  mutations_info %>% 
  select(primary_disease, primary_or_metastasis) %>% 
  left_join(mutations_per_cancer, by = "primary_disease") 
```

Plot distribution of cancer types across cell lines
```{r, fig.width=5, fig.height=3}
ggplot(data=plot_CCLE) +
  geom_col(mapping = aes(x = reorder(primary_disease, freq),y = freq, fill = primary_or_metastasis)) +
  xlab(NULL) + 
  ylab(NULL) + 
  labs(fill = "Primary or Metastasis") + 
  ggtitle("Number of primary and metastatic cancer in each cell line") +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  scale_fill_manual(values = wes_palette("GrandBudapest1", n = 2, type = "discrete"),na.value ="black") +
  coord_flip() 
  #(axis.text.x = element_text(angle = 90)) +       #rotate the label
 
```

Repeat the plot above with counts normalized by number of cell lines studied
```{r}
##MUTATIONS PER CELL LINE
cell_lines_studied = 
  mutations_info %>% 
  group_by(primary_disease) %>% 
  count(CCLE_Name) %>% 
  dplyr::rename("observed_mutations_percelline" = "n")

## STUDIED CELL LINES PER CANCER
normalize = 
  cell_lines_studied %>% 
  count(primary_disease) %>%
  rename("cell_lines_number" = "n") 

## MUTATIONS ACROSS ALL CELL LINES
cell_lines_distribution = 
  cell_lines_studied %>% 
  count(primary_disease, wt = observed_mutations_percelline) %>% 
  rename("observed_mutation_all_cell_lines" = "n") 


```


Show distribution of cell lines across the different cancer types (how many cell lines have been studied per cancer)
```{r, fig.height=3, fig.width=4}
ggplot(data = normalize, aes(x = reorder(primary_disease, cell_lines_number), y = cell_lines_number, fill="black")) +
  geom_bar(stat = "identity", fill = "black")+ 
  xlab(NULL) + 
  ylab("Number of cell lines studied")+ 
  ggtitle("Number of cell lines studied per cancer") +
  geom_text(aes(label = cell_lines_number), size = 3, position = "dodge", hjust=-0.2) +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  coord_flip() 
```

Try to do an average of mutations (mean of mutations)
```{r}
cell_lines_distribution = 
  right_join(cell_lines_distribution, normalize, by = "primary_disease", copy = TRUE)

cell_lines_distribution = 
  cell_lines_distribution %>% 
  mutate(normalized_mutations = observed_mutation_all_cell_lines / cell_lines_number) %>% 
  arrange(normalized_mutations)

ggplot(data = cell_lines_distribution, aes(x = reorder(primary_disease, normalized_mutations), y = normalized_mutations)) +
  geom_col(fill = "black") +
  geom_text(aes(label = sprintf("%0.0f", round(normalized_mutations, digits =0))), size = 2.5, position = "dodge", hjust=-0.05) +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  coord_flip() + 
  xlab(NULL) + 
  ylab("Observed mutations normalized for cell lines studied")

```

Retrieve database with HUGO symbol nomenclature 
```{r}
#ensembl <- useMart("ensembl")
#datasets <- listDatasets(ensembl)
#view(datasets)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl") 

#get list of all available filters
#filters <- listFilters(ensembl) 

```

## Get gene length
Fetch gene symbols, to be used instead of gene ensembl id
```{r}
#create vector with the Hugo Symbols that we want
id_list = mutations_info %>% 
  select(Hugo_Symbol) %>% 
  distinct() %>% 
  pull(Hugo_Symbol)

#gene list with hugo symbol and entrez id / ensembl id 
genes_list =
 getBM(filters= "hgnc_symbol", attributes= c("ensembl_gene_id","hgnc_symbol", "description"), 
       values = id_list, uniqueRows = TRUE, mart = useDataset("hsapiens_gene_ensembl", useMart("ensembl")))

ensembl_list =  
  genes_list %>% 
  pull(ensembl_gene_id)

#GET GENE AND TRANSCRIPT LENGTH
getlength(ensembl_list, 'hg19', 'ensGene') #output into hg19.ensGene.LENGTH

#rename length output
length = hg19.ensGene.LENGTH
#write.csv(length, "./localdata/gene_length.csv") #save as csv to local 
length = 
  length %>% 
  rename("ensembl_gene_id" = "Gene") #obs length depends on transcript id, no gene id

#merge length and gc content with hugo symbol & gene info
gene_length_list = 
  left_join(genes_list, length, by = "ensembl_gene_id", copy = TRUE) %>%
  select(hgnc_symbol, Length, Transcript) %>% 
  rename("Hugo_Symbol" = "hgnc_symbol")

write_csv(gene_length_list, "../../../localdata/gene_length.csv", quote = FALSE)
```

## Normalize for gene length

Normalize the amount of mutations by length of the gene
```{r}
#GET ENS ID OF AFFECTED TRANSCRIPT
transcripts =
mutations_info %>% 
  select(DepMap_ID, Hugo_Symbol, Annotation_Transcript)
  
transcripts$Annotation_Transcript = gsub("\\..*", "", transcripts$Annotation_Transcript) 


#FIND DUPLICATES
test_tr_1 = 
transcripts %>%  
  filter(DepMap_ID == "ACH-000001")  

test_tr_1[test_tr_1$Hugo_Symbol %in% test_tr_1$Hugo_Symbol[duplicated(test_tr_1$Hugo_Symbol)],]
  
```




```{r}
#count mutations per gene and cancer, add transcript info

##GET RID OF TRANSCRIPT VERSION (ENSXXXXX.X)
mutations_info$Annotation_Transcript = gsub("\\..*", "", mutations_info$Annotation_Transcript) 

  
mut_per_gene_cancer = 
  mutations_info %>% 
  group_by(Annotation_Transcript, primary_disease) %>% 
  tally() %>% 
  #count w/o adding extra grouping 
  rename("mutation_nr" = "n") %>% 
  left_join(transcripts, by = "Annotation_Transcript") %>% 
  rename("Transcript" = "Annotation_Transcript") %>% 
  ungroup() %>% 
  select(-Hugo_Symbol) 

#normalize per length of transcript
norm_mut_per_gene_cancer = 
left_join(mut_per_gene_cancer, gene_length_list, by="Transcript") %>% 
  mutate(norm_mutations = mutation_nr/Length) 
```

## Top mutated genes per cancer

```{r}
##CREATE PALETTE FOR PLOT 
#DEFINE NUMBER OF COLORS
nb.cols <- 50
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(nb.cols)

top_mutations = 
norm_mut_per_gene_cancer %>% 
  select(-DepMap_ID) %>% 
  arrange(desc(norm_mutations)) %>%
  group_by(primary_disease) %>% 
  unique() %>% 
  top_n(3) 

top_mutations = 
  top_mutations %>% 
  filter(Hugo_Symbol != "NA")

top_mutations$primary_disease = 
reorder(top_mutations$primary_disease, top_mutations$norm_mutations)

ggplot(data = top_mutations, aes(x = reorder(primary_disease, norm_mutations), y = norm_mutations, fill = Hugo_Symbol)) +
  geom_col() + 
  scale_fill_manual(values = mycolors) +
  #geom_text(aes(label = Hugo_Symbol)) +
  coord_flip() + 
  xlab(NULL) + 
  ylab("Most mutated genes per cancer") + 
  theme_bw() + 
  theme(
  panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())
```


## Top mutated genes across all cancers (considering only most mutated transcript)
```{r}
most_mutated_genes = 
  mutations_info %>% 
  filter(Variant_Classification != "Silent") %>% 
  group_by(Hugo_Symbol) %>% 
  tally() %>% 
  rename("observed_mutations" = "n") %>% 
  left_join(gene_length_list, by = "Hugo_Symbol") %>% 
  mutate(normalized_mutations = observed_mutations/Length) %>% 
  arrange(desc(normalized_mutations)) %>%
  distinct(Hugo_Symbol ,.keep_all = TRUE) %>% #discard alternative transcripts, keep most mutated one
  top_n(50)

#reorder
most_mutated_genes$Hugo_Symbol <- reorder(most_mutated_genes$Hugo_Symbol, most_mutated_genes$normalized_mutations)
```

plot
```{r, fig.width=5, fig.height=6}
  ggplot(data = most_mutated_genes, aes(x = Hugo_Symbol, y = normalized_mutations)) +
  geom_col(fill = "black") +
  coord_flip() +
  theme_bw()+ 
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  xlab(NULL) +
  ylab("Most mutated genes across cancer cell lines")

```

# Barplot with mutation classification
```{r}
##AVOID SCIENTIFIC NOTATION
options(scipen = 1000000)

##COUNT OCCURENCE OF EACH MUTATION CLASS
mutation_class = 
  CCLE_mutations %>% 
  count(Variant_Classification) %>% 
  rename("Frequency" = "n")

##PLOT
ggplot(data = mutation_class, aes(x = reorder(Variant_Classification, Frequency), y = Frequency, fill="black")) +
  geom_bar(stat = "identity", fill = "black")+ 
  xlab("Mutation classification") + 
  ylab("Number of mutations")+ 
  ggtitle("") +
  theme_bw() +
  theme(axis.ticks.x = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  coord_flip()
```

