---
title: "Intersection plot - human protein atlas spliceosome list"
output: html_notebook
---

Description: intersection plot of cell lines and mutated spliceosome genes (TCGA list). 

```{r, message=FALSE, include=FALSE}
library(tidyverse)
library(readr)
library(biomaRt)
library(car)
library(cowplot)
library(ggpubr)
library(ggsci)
library(ggrepel)
library(ggExtra)
library(hrbrthemes)
library(wesanderson)
library(data.table)
library(biomaRt)
library(KEGGREST)
library(plyr)
library(UpSetR)
library(RColorBrewer)
library(ComplexHeatmap)
library(tidyHeatmap)
select = dplyr::select
rename = dplyr::rename
count = dplyr::count
```

```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")
```

Load data   
```{r}
CCLE_mutations = fread("depmap/ccle_info_21q1.csv", header = TRUE) #mutations and sample info 

#core spliceosome
core_spliceosome = fread("spliceosome/core_hp.tsv", sep = "\t", header = TRUE)

#SELECT VARIABLES OF INTEREST
core_spliceosome = 
  core_spliceosome %>% 
  select(Gene) %>% 
  rename("Hugo_Symbol" = "Gene")
```


## Identify spliceosome mutated genes in CCLE
Filter CCLE using the generated gene list. 
```{r}
## FILTER MUTATIONS THAT AFFECT THE SPLICEOSOME
core_spliceosome_mutations_hp = 
   #include genes present in CCLE and in KEGG pathway: only genes related to the spliceosome
  inner_join(CCLE_mutations, core_spliceosome, by = "Hugo_Symbol") %>%
  #select variables of interest
  dplyr::select(Hugo_Symbol, Variant_Classification, Variant_Type, stripped_cell_line_name, primary_or_metastasis, 
                primary_disease, Subtype, Variant_annotation, DepMap_ID, Annotation_Transcript) %>% 
  filter(Variant_Classification != "Silent" | Variant_Type != "SNP")  

```

## Normalization
Get nr of cell lines studied per disease
```{r}
## DF WITH NUMBER OF STUDIED CELL LINES PER CANCER
normalize_cell_lines = 
  CCLE_mutations %>% 
  group_by(primary_disease) %>% 
  dplyr::distinct(stripped_cell_line_name) %>% 
  #count n of distinct cell lines for each cancer #sometimes problems with masking of function
  dplyr::count(primary_disease) %>% 
  rename("cell_lines_number" = "n") 
```

Average number of mutations per cancer
```{r}
core_normalized_mutations_cancer =   
core_spliceosome_mutations_hp %>% 
  group_by(primary_disease) %>% 
  #count observations per cancer  
  tally() %>% 
  rename("MUTATIONS_PER_CANCER" = "n") %>% 
  left_join(normalize_cell_lines, by = "primary_disease") %>% 
  #normalize for the number of cell lines studied
  mutate(NORMALIZED_MUTATIONS = MUTATIONS_PER_CANCER / cell_lines_number) %>% 
  rename("DISEASE" = "primary_disease")
```

Plot
```{r, fig.height=2.5, fig.width4}
  ggplot(data=core_normalized_mutations_cancer, aes(x = reorder(DISEASE, NORMALIZED_MUTATIONS), y = NORMALIZED_MUTATIONS)) +
  geom_bar(stat = "identity", fill = "black") +
  geom_text(aes(label = sprintf("%0.2f", round(NORMALIZED_MUTATIONS, digits =2))), size = 2, position = "dodge", hjust=-0.1) +
  xlab(NULL) + 
  ylab("Average number of mutations per cell line") + 
  ggtitle("Number of mutations in the core splicing machinery") +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "none",
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  coord_flip() 

#ggsave("../figures/exploration/spliceosome/nr_mutations_core_spliceosome_percancer.png", height = 5, width = 7)
```

Annotate mutations to get a binary matrix for upset plot 
```{r}
core_spliceosome_mutations_hp = 
  core_spliceosome_mutations_hp %>% 
  mutate(mutated = 1) %>% 
  distinct()
```


Pivot wide core_spliceosome_mutations
```{r}
matrix_core = 
  core_spliceosome_mutations_hp %>% 
  pivot_wider(id_cols = DepMap_ID, names_from = Hugo_Symbol, values_from = mutated, values_fill = 0, values_fn = length)

dim(matrix_core)
```

Let's do upset plot per cancer
```{r}
core_spliceosome_mutations = 
  core_spliceosome_mutations_hp %>% 
  mutate(mutated = 1) %>% 
  distinct() 
```

```{r}
matrix_core = 
  core_spliceosome_mutations%>% 
  pivot_wider(id_cols = DepMap_ID, names_from = Hugo_Symbol, values_from = mutated, values_fill = 0, values_fn = length) 


#PIVOT LONG FOR EASY MUTATION
matrix_long = 
  matrix_core %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "Genes", values_to = "mutated") %>% 
  mutate(mutated = case_when(
    mutated == 0 ~ 0, 
    mutated != 0 ~ 1
  ))

#BACK TO WIDE
matrix_core = 
  matrix_long %>% 
  pivot_wider(id_cols = DepMap_ID, names_from = Genes, values_from = mutated) %>% 
  as.data.frame()

```

```{r}
sets = 
  matrix_long %>% 
  select(Genes) %>% 
  unique() %>% 
  as_vector()

```

```{r, fig.height=7, fig.width=7}
upset(matrix_core, sets = sets_endo, order.by = "freq", empty.intersections = "off", point.size = 3,)

 #c("NONO", "CCAR1", "POLR2B", "DHX9", "PCF11", "POLR2I", "POLR2A", "SNRNP200", "RBMX", "SON", "POLR2C", "GEMIN4"
```

