---
title: "Analysis of the spliceosome mutations, using tidyHeatmap"
output: html_notebook
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis")
```

```{r, message=FALSE, include=FALSE}
library(tidyverse)
library(readr)
library(biomaRt)
library(car)
library(cowplot)
library(ggpubr)
library(ggsci)
library(ggrepel)
library(ggExtra)
library(hrbrthemes)
library(wesanderson)
library(data.table)
library(biomaRt)
library(KEGGREST)
library(plyr)
library(UpSetR)
library(RColorBrewer)
library(ComplexHeatmap)
library(tidyHeatmap)
library(tidyheatmap)
select = dplyr::select
rename = dplyr::rename
count = dplyr::count
```
Renviron
source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")
```


## Load data

Read data: list of mutations from CCLE and list of genes that are part of the spliceosome machinery
```{r,  message=FALSE}
##SNPs
CCLE_mutations = fread("localdata/depmap/mutations_info", header = TRUE) #mutations and sample info 

##FROM HGCN, MAJOR SPLICEOSOME GENES
splicing_major_genes_list = fread("/Users/castilln/Desktop/thesis/localdata/spliceosome/group-major-spliceosome.csv", header = TRUE) 

splicing_major_genes_list = 
splicing_major_genes_list %>% 
 dplyr::rename("Hugo_Symbol" = "Approved symbol") 

##FROM HGCN, MINOR SPLICEOSOME GENES
splicing_minor_genes_list = fread("/Users/castilln/Desktop/thesis/localdata/spliceosome/group-minor-spliceosome.csv", header = TRUE) 
splicing_minor_genes_list = 
splicing_minor_genes_list %>% 
   rename("Hugo_Symbol" = "Approved symbol") 

##MAKE LIST WITH ALL GENES ASSOCIATED WITH THE SPLICEOSOME
splicing_gene_list = 
  full_join(splicing_major_genes_list, splicing_minor_genes_list) %>% 
  select(Hugo_Symbol)

##LOAD SPLICEOSOME ASSOCIATED GENES FROM HUMAN PROTEIN ATLAS
spliceosome_hp = fread("/Users/castilln/Desktop/thesis/localdata/spliceosome/spliceosome.tsv", sep = "\t", header = TRUE) 

#SELECT VARIABLES OF INTEREST
spliceosome_hp = 
  spliceosome_hp %>% 
  select(Gene) %>% 
  rename("Hugo_Symbol" = "Gene")
```

Gene length for normalization
```{r}
gene_length = fread("localdata/gene_length.csv", select = c("Hugo_Symbol", "Transcript", "Length"))
```


## Prepare data frame with mutations affecting the splicing machinery / process
Query KEGG and obtain hugo symbols for the genes belonging to spliceosome pathway
```{r}
##GET SPLICEOSOMAL PATHWAT FROM KEGG
query = keggGet(c("hsa03040")) 

##OBTAIN GENE LIST BELONGING TO PATHWAY
res = keggLink("hsa", "path:hsa03040")
res = as.list(res)

##EMPTY VECTOR WITH SAME LENGTH AS NR OF GENES IN PATHWAY 
ncbi_id = vector(mode = "character", length = length(res)) 

## CREATE LIST WITH GENE NAMES
for (i in 1:length(res)){
  #access each gene individually
  gene = res[i] 
  #get NCBI ID for each gene
  ncbi_id[i] = keggConv("ncbi-geneid", gene) 
}

##ONLY ID
ncbi_id = as.vector(str_remove(ncbi_id, "ncbi-geneid:")) 

## LIST OF HUGO SYMBOLS FOR GENES RELATED TO SPLICEOSOME ACCORDING TO KEGG
genes_list = getBM(filters= "entrezgene_id", attributes= c("entrezgene_id",
"hgnc_symbol", "description"), values = ncbi_id, uniqueRows = TRUE, mart = useDataset("hsapiens_gene_ensembl", useMart("ensembl"))) 
```

```{r, include=FALSE}
## SAVE LIST AS .CSV
#write_csv(genes_list, "../../../localdata/depmap/gene_list_from_kegg.csv")

#genes_list = fread("../../../localdata/depmap/gene_list_from_kegg.csv")
```

## Identify spliceosome mutated genes in CCLE

Filter CCLE using the generated gene list. 
```{r}
##RENAME ID TO HUGO SYMBOL
genes_list = 
  genes_list %>% 
  rename("Hugo_Symbol" = "hgnc_symbol") %>% 
  select(Hugo_Symbol)

##BIND GENE LIST FROM THE DIFFERENT SOURCES
spliceosome_genes = 
  rbind(spliceosome_hp, splicing_gene_list, genes_list) %>% 
  unique()

## FILTER MUTATIONS THAT AFFECT THE SPLICEOSOME
spliceosome_mutations_ccle = 
   #include genes present in CCLE and in KEGG pathway: only genes related to the spliceosome
  inner_join(CCLE_mutations, spliceosome_genes, by = "Hugo_Symbol") %>%
  #select variables of interest
  dplyr::select(Hugo_Symbol, Variant_Classification, Variant_Type, stripped_cell_line_name, primary_or_metastasis, 
                primary_disease, Subtype, Variant_annotation, DepMap_ID, Annotation_Transcript) %>% 
  filter(Variant_Classification != "Silent" | Variant_Type != "SNP")  

```


```{r}
write_csv(spliceosome_mutations_ccle, "localdata/depmap/mutations_spliceosome.csv")

```


## Normalization

Get nr of cell lines studied per disease
```{r}
## DF WITH NUMBER OF STUDIED CELL LINES PER CANCER
normalize_cell_lines = 
  CCLE_mutations %>% 
  group_by(primary_disease) %>% 
  dplyr::distinct(stripped_cell_line_name) %>% 
  #count n of distinct cell lines for each cancer #sometimes problems with masking of function
  dplyr::count(primary_disease) %>% 
  rename("cell_lines_number" = "n") 
```

Average number of mutations per cancer
```{r}
normalized_mutations_cancer =   
spliceosome_mutations_ccle %>% 
  group_by(primary_disease) %>% 
  #count observations per cancer  
  tally() %>% 
  rename("MUTATIONS_PER_CANCER" = "n") %>% 
  left_join(normalize_cell_lines, by = "primary_disease") %>% 
  #normalize for the number of cell lines studied
  mutate(NORMALIZED_MUTATIONS = MUTATIONS_PER_CANCER / cell_lines_number) %>% 
  rename("DISEASE" = "primary_disease")
```

Plot
```{r, fig.height=2.5, fig.width=3}
  ggplot(data=normalized_mutations_cancer, aes(x = reorder(DISEASE, NORMALIZED_MUTATIONS), y = NORMALIZED_MUTATIONS)) +
  geom_bar(stat = "identity", fill = "black") +
  geom_text(aes(label = sprintf("%0.2f", round(NORMALIZED_MUTATIONS, digits =2))), size = 2, position = "dodge", hjust=-0.1) +
  xlab(NULL) + 
  ylab("Average number of mutations per cell line") + 
  ggtitle("Number of mutations in the splicing machinery") +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "none",
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  coord_flip() 
```

Normalize mutations per transcript length
```{r}
normalized_mutations_gene =
  spliceosome_mutations_ccle %>%
  group_by(Hugo_Symbol, Annotation_Transcript) %>% 
  tally() %>% 
  rename("MUTATIONS" = "n") 

normalized_mutations_gene$Annotation_Transcript =   
  gsub("\\..*", "", normalized_mutations_gene$Annotation_Transcript)

gene_length = 
  gene_length %>% 
  select(-Hugo_Symbol)

normalized_mutations_gene = 
  normalized_mutations_gene %>%   
  rename("Transcript" = "Annotation_Transcript") %>% 
  left_join(gene_length, by = "Transcript") %>% 
  #normalize by gene length
  mutate(NORMALIZED_MUTATIONS = MUTATIONS / Length) %>% 
  arrange(desc(NORMALIZED_MUTATIONS)) %>% 
  ungroup()

normalized_mutations_gene %>% 
  top_n(20)
```

Most mutated spliceosomal genes per cancer 
```{r}
top_mutations_cancer = 
spliceosome_mutations_ccle %>% 
  select(Hugo_Symbol, primary_disease, Annotation_Transcript, stripped_cell_line_name) %>% 
  rename("Transcript" = "Annotation_Transcript") %>% 
  group_by(Hugo_Symbol, Transcript) %>%
  count(primary_disease) %>% 
  rename("MUTATIONS" = "n") %>% 
  ungroup() %>% 
  group_by(primary_disease) %>% 
  arrange(desc(MUTATIONS)) %>% 
  top_n(3)


##FOR VISUALIZATION: NORMALIZE PER TRANSCRIPT LENGTH 
#THERE IS NO LENGTH INFORMATION FOR ALL GENES
norm_top_mutations_cancer = 
  top_mutations_cancer %>% 
  left_join(gene_length, by ="Transcript") %>% 
  mutate(NORM_MUTATIONS = MUTATIONS/Length)
```

```{r, fig.height=20, fig.width=4}
colorCount = length(unique(top_mutations_cancer$Hugo_Symbol))
getPalette = colorRampPalette(brewer.pal(9, "Paired"))

ggplot(top_mutations_cancer, aes(x = Hugo_Symbol, y = MUTATIONS, group_by(primary_disease))) +
  geom_col() +
  facet_wrap("primary_disease") +
  scale_fill_manual(values = getPalette(colorCount)) +
  geom_text_repel(aes(x = primary_disease, y = MUTATIONS, label = Hugo_Symbol), size = 1, 
                  segment.color = '#cccccc',     segment.size = 0.2,
colour = "black", point.padding = NA, position = position_identity()) + 
  coord_flip() +
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
   theme(legend.position="bottom") +
  guides(fill=guide_legend(nrow=6))+
  xlab(NULL) +
  ylab("Most mutated genes across cancer types")
```

Heatmap
```{r}
##PREPARE FOR HEATMAP
#PIVOT WIDER
df = 
  top_mutations_cancer %>%
  select(Hugo_Symbol,primary_disease, MUTATIONS) %>% 
  pivot_wider(id_cols = NULL, names_from = primary_disease, values_from = MUTATIONS)%>% 
  filter(!is.na(Hugo_Symbol))  #pivot wider to get a df with mutations, disease and relative mutation rate

df[is.na(df)] <- 0 #convert na values in 0

df %>%
  select(-c(Engineered, Unknown)) %>% 
  rename("Gene"="Hugo_Symbol") -> df #remove Engineered and Unknown classes

df[ , colSums(is.na(df)) == 0] -> df #drop columns containing na values

df %>% 
  pivot_longer(cols = !Gene,  names_to = "DISEASE", values_to = "Relative mutation") %>% 
  left_join(normalized_mutations_cancer, by = "DISEASE") -> df_hm_tidy

#tidy heatmap
#heatmap of number of mutations for each gene across cancer: only non-silent and non-SNP mutations
df_hm_tidy %>%
  heatmap(Gene, DISEASE, `Relative mutation`,
          palette_value = circlize::colorRamp2(c(0, 2, 5), viridis::magma(3))) %>% 
  add_bar(NORMALIZED_MUTATIONS) %>% 
  add_bar(cell_lines_number) -> htm
##transform relative mutations in percentage
htm
```
analysis for individual cell lines
```{r}
cell_line_spliceosome_mutations_ccle <-spliceosome_mutations_ccle %>%
  group_by(stripped_cell_line_name) %>% #group by cell line
  dplyr::count(Hugo_Symbol) %>% #count nr of mutations for each gene/cell_line
  rename("observed_mutations" = "n")  
```

Prepare for heatmap
pivot wider
```{r}
df_cell_lines <- cell_line_spliceosome_mutations_ccle %>% 
  ungroup() %>% 
  pivot_wider(id_cols = Hugo_Symbol, names_from = stripped_cell_line_name, values_from = observed_mutations)%>% 
  filter(!is.na(Hugo_Symbol))  #pivot wider to get a df with mutations, disease and relative mutation rate

df_cell_lines[is.na(df_cell_lines)] <- 0 #convert na values in 0

```

tidy heatmap
```{r}
#heatmap with amount of mutations for each spliceosomal gene across cell lines
df_cell_lines %>%
  rename("Gene" = "Hugo_Symbol") -> df_cell_lines #remove Engineered and Unknown classes

df_cell_lines[ , colSums(is.na(df_cell_lines)) == 0] -> df_cell_lines #drop columns containing na values
  
df_cell_lines %>% 
  pivot_longer(cols = !Gene,  names_to = "Cell line", values_to = "Relative mutation") -> df_tidy

#heatmap
library(tidyHeatmap)
df_tidy %>%
  heatmap(Gene, `Cell line`, `Relative mutation`,
           palette_value = circlize::colorRamp2(c(0, 5, 10, 15, 20), viridis::magma(5))) -> htm_cell_lines

htm_cell_lines

```

repeat heatmap only for damaging mutations (according to variant_annotation variable)
```{r}
spliceosome_mutations_ccle_cell_line_dam =
  spliceosome_mutations_ccle %>%
  filter(Variant_annotation == "damaging") %>% 
  group_by(stripped_cell_line_name) %>% #group by cell line
  dplyr::count(Hugo_Symbol) %>% #count nr of mutations for each gene/cell_line
  rename("observed_mutations" = "n")   

df_cell_lines_dam = 
  spliceosome_mutations_ccle_cell_line_dam %>% 
  ungroup() %>% 
  pivot_wider(id_cols = Hugo_Symbol, names_from = stripped_cell_line_name, values_from = observed_mutations)%>% 
  filter(!is.na(Hugo_Symbol))   #pivot wider to get a df with mutations, disease and relative mutation rate


df_cell_lines_dam[is.na(df_cell_lines_dam)] <- 0 #convert na values in 0

#heatmap with amount of mutations for each spliceosomal gene across cell lines
df_cell_lines_dam = 
  df_cell_lines_dam %>%
  rename("Gene" ="Hugo_Symbol")  

df_cell_lines_dam[ , colSums(is.na(df_cell_lines_dam)) == 0] -> df_cell_lines_dam #drop columns containing na values

df_tidy_dam = 
  df_cell_lines_dam %>% 
  pivot_longer(cols = !Gene,  names_to = "Cell line", values_to = "Relative mutation") 

#heatmap
htm_cell_lines_dam = 
  df_tidy_dam %>%
  heatmap(Gene, `Cell line`, `Relative mutation`,
          palette_value = circlize::colorRamp2(c(0, 5, 10, 15, 20), viridis::magma(5))) 

htm_cell_lines_dam

```

