---
title: "Exploration of CLL dataset"
output: html_notebook
---

load libraries
```{r, message=FALSE, include=FALSE}
library(tidyverse)
library(readr)
library(biomaRt)
select = dplyr::select
```

load data
```{r, message=FALSE, include=FALSE}
CCLE_mutations <- read_delim("../data/CCLE_mutations.csv", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
#head(CCLE_mutations)

sample_info <- read_csv("../data/sample_info.csv")
#head(sample_info)
```

merge both datasets (by DepMap_ID)
```{r}
joined_CCLE <- right_join(CCLE_mutations, sample_info, by = "DepMap_ID", copy = TRUE, suffix = c(".CCL", ".info"))
```

Distribution of cancer types across cell lines
```{r}
ggplot(data=joined_CCLE) +
  geom_bar(mapping = aes(x = primary_disease, fill = primary_or_metastasis)) +
  coord_flip() + 
  xlab(NULL) + 
  labs(fill = "Primary or Metastasis")
  #(axis.text.x = element_text(angle = 90)) +       #rotate the label
 
```

Repeat the plot above with counts normalized by number of cell lines studied
```{r}
cell_lines_studied <- joined_CCLE %>% 
  group_by(primary_disease) %>% count(CCLE_Name) %>% rename(observed_mutations = n)

#count the number of cell lines per cancer type 
normalize <- cell_lines_studied %>% 
  count(primary_disease) %>% rename(cell_lines_number = n)

cell_lines_distribution <- cell_lines_studied %>% 
  count(primary_disease, wt = observed_mutations) %>% rename(observed_mutation_all_cell_lines = n) 



```

Show distribution of cell lines across the different cancer types (how many cell lines have been studied per cancer)
```{r}
ggplot(data = cell_lines_studied) +
  geom_bar(mapping = aes(x = primary_disease)) + 
  coord_flip() + 
  xlab(NULL) + 
  ylab("Number of cell lines studied") 
   
```


```{r}
cell_lines_distribution <- right_join(cell_lines_distribution, normalize, by = "primary_disease", copy = TRUE)

cell_lines_distribution %>% 
  mutate(normalized_mutations = observed_mutation_all_cell_lines / cell_lines_number) -> cell_lines_distribution

ggplot(data = cell_lines_distribution) +
  geom_col(mapping = aes(x = primary_disease, y = normalized_mutations)) + 
  coord_flip() + 
  xlab(NULL) + 
  ylab("Observed mutations normalized for cell lines studied")
```

Subset pancreatic cancer, cell lines used
```{r}

joined_CCLE %>% 
  filter(primary_disease == "Pancreatic Cancer") -> pancreatic_cancer

head(pancreatic_cancer)
```

Retrieve database with HUGO symbol nomenclature 
```{r}
#ensembl <- useMart("ensembl")
#datasets <- listDatasets(ensembl)
#view(datasets)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl") 

#get list of all available filters
#filters <- listFilters(ensembl) 

```


Fetch gene symbols, to be used instead of gene ensembl id
```{r}
#create vector with the Hugo Symbols that we want
id_list = joined_CCLE %>% 
  select(Hugo_Symbol) %>% 
  distinct() %>% 
  pull(Hugo_Symbol)

genes_list <- getBM(filters= "hgnc_symbol", attributes= c("ensembl_gene_id",
"hgnc_symbol", "description"), values = id_list, uniqueRows = TRUE, mart = useDataset("hsapiens_gene_ensembl", useMart("ensembl")))
```

