---
title: "Does any of the SpliceAI predicted splicing mutations have a high dependency?"
output: html_notebook
---

Load libraries 
```{r}
library(tidyverse)
library(data.table)
```

Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis")
```

Load linear model & df
```{r}
##DEPENDENCIES
dependencies_df <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_gene_dependency.csv", header = TRUE)

#MUTATIONS AND METADATA
CCLE_mutations <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_info", header = TRUE) 


## READ SPLICEAI RESULTS
predicted_splice <- readRDS("github/results/spliceAI/ths8.rds") 
```

From spliceAI predictions, select the off-target mutations
```{r}
predicted_na = 
  predicted_splice %>% 
  filter(is.na(DepMap_ID)) %>% 
  select(-c("primary_disease", "Variant_Classification", "Annotation_Transcript"))
```

Annotate primary disease
```{r}
head(CCLE_mutations)

lines_dis = 
  CCLE_mutations %>% 
  select(primary_disease, stripped_cell_line_name, DepMap_ID) %>% 
  distinct() %>% 
  rename("ID" = "stripped_cell_line_name",
         "DISEASE" = "primary_disease")
predicted_na = 
  predicted_na %>% 
  left_join(lines_dis, by = "ID") %>% 
  as.data.table()
```
Create a unique ID for each SNP so that we can query the CCLE df
```{r}
##UNIQUE VARIABLE FOR SPLICEAI OUT
id_na = predicted_na[, id_var := paste(ID, SYMBOL, sep = "_")]

id_ccle = CCLE_mutations[, id_var := paste(stripped_cell_line_name, Hugo_Symbol, sep = "_"), by = .(stripped_cell_line_name, Hugo_Symbol)]
```

Filter CCLE using the off-target predicted mutations
```{r}
na_snps = 
  semi_join(id_ccle, id_na, by = "id_var")
```

In this way we get a DepMap ID for the SNP, but keep in mind that that DepMap ID refers to other mutation in the same gene. 

Extract the DepMap ID from the genes predicted by SpliceAI to suffer off target mutations
```{r}
depmap_id = 
  na_snps %>% 
  pull(DepMap_ID) %>% 
  unique() %>% 
  as.data.frame() 

colnames(depmap_id) <- "DepMap_ID"
```

Use the DepMap ID to filter those cell lines for which dependencies have been calculated that also had predicted off target splicing
```{r}
filt_dependencies = 
  dependencies_df %>% 
  semi_join(depmap_id, by = "DepMap_ID")
```

Pivot longer & create a new column for entrez ID 
```{r}
long_dependencies = 
  filt_dependencies %>% 
  pivot_longer(cols = -DepMap_ID, names_to = "SYMBOL", values_to = "DEPENDENCY" )

long_dependencies = 
  long_dependencies %>%
    separate(col = "SYMBOL", into = c("SYMBOL", "ENTREZ"), sep = "\\s") 
   
long_dependencies$ENTREZ <- gsub("\\(|\\)","",as.character(long_dependencies$ENTREZ))
```

Find the dependency for the specific gene
```{r}
long_dependencies =
  long_dependencies %>% 
  select(-ENTREZ)

pred_symbol_depmap = 
  na_snps %>% 
  select(DepMap_ID, Hugo_Symbol) %>% 
  distinct() %>% 
  rename("SYMBOL" = "Hugo_Symbol")

long_dependencies =
  long_dependencies %>% 
  semi_join(pred_symbol_depmap, by = c("DepMap_ID", "SYMBOL"))

long_dependencies %>% 
  arrange(desc(DEPENDENCY)) %>% head()
```

```{r}
head(id_na)

id_na %>% 
  filter(SYMBOL == "SMC4")
```

