---
title: "MSK impact gene effect"
output: html_notebook
---

Description: list of off-target mutations (mutations in one position that modify an splice site upstream or downstream of that particular position). 

Libraries
```{r}
library(tidyverse)
library(data.table)
library(readr)


select = dplyr::select
rename = dplyr::rename
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Set wd 
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```

Load data
```{r}
#DEPMAP somatic mutations
ccle = fread("depmap/ccle_info_21q1.csv")
ccle = 
  ccle %>% 
  rename("SYMBOL" = "Hugo_Symbol")

#CRISPR gene effect
gene_effect = fread("depmap/achilles/Achilles_gene_effect.csv")

#LOAD ACTIONABLE SPLICE OUTPUT (ALREADY FILTERED FOR MSK) 
actionable_splice = readRDS("msk-impact/spliceOut_actionable.rds")
head(actionable_splice)
```

Pivot longer gene effect 
```{r}
df_long_crispr = 
  gene_effect %>% 
  pivot_longer(cols = -DepMap_ID, names_to = "SYMBOL", values_to = "gene_effect") %>% 
  mutate(SYMBOL = gsub(" .*","",SYMBOL))
```


```{r}
ccle = 
  ccle %>% 
  select(-V1)

head(ccle)

ccle_join = 
  ccle %>%  
  select(SYMBOL, DepMap_ID, Variant_Type, isDeleterious)
```
Join gene effect with metadata
```{r}
df_full = 
  full_join(df_long_crispr, ccle_join, by = c("SYMBOL", "DepMap_ID")) %>% 
  filter(!is.na(gene_effect)) %>% 
  distinct()

# Some of the genes do not have mutations - substitute NA for WT information
df_full = 
  df_full %>% 
  mutate(Variant_Type = ifelse(is.na(Variant_Type), "WT", Variant_Type),
         isDeleterious = ifelse(is.na(isDeleterious), F, isDeleterious))


head(df_full)
```
Filter only those genes in MSK 
```{r}
# List of actionable genes
act_list = 
  actionable_splice %>% 
  select(SYMBOL) 

act_depmap = 
  actionable_splice %>% 
  select(DepMap_ID, SYMBOL)

# Get gene effect of the variants that cause splicing 
df_full_act = 
  df_full %>% 
  inner_join(act_list, by = "SYMBOL") %>% 
  distinct()


head(df_full_act)

# Use SpliceAI output to select those mutations+cell lines predicted to be splice mutating by SpliceAI
df_labelled = 
  df_full_act %>%
  inner_join(act_depmap, by = c("SYMBOL", "DepMap_ID")) %>% 
  mutate(spliceAI = T)
```
Join back with rest of gene effect info
```{r}
# Join with rest of data
df_all = 
  df_labelled %>% 
  full_join(df_full_act) %>% 
  distinct()

# SpliceAI false for those that were not predicted 
df_all = 
  df_all %>% 
  mutate(spliceAI = ifelse(is.na(spliceAI), F, spliceAI))
```

Calculate proportions of spliceAI predicted mutations and rest 
```{r}
head(df_all)

table(df_all$spliceAI)

#0.2% of events are SpliceAI positive 
```

```{r}
library(purrr)
library(ggpubr)
library(ggstatsplot)
#Try subsampling for easier plotting
set.seed(1337)

nested_df <- df_all %>% 
  group_by(spliceAI) %>% 
  nest() %>% 
  ungroup() %>% 
  #select from True and False to keep the proportion from the large df 
  mutate(n = c(677, 20000))


sampled_df <- nested_df %>%
  mutate(samp = map2(data, n, sample_n))

small_df = 
  sampled_df %>% 
  select(-c(data, n)) %>% 
  unnest()
```


```{r}
small_df %>%  
  ggbetweenstats(
    x = spliceAI, 
    y = gene_effect, 
    p.adjust.method = "bonferroni",
    #scale_fill_manual(values = c("gray", "red")),   
    xlab = "Predicted alteration by SpliceAI", 
    ylab = "Gene effect",
    title = "Comparison of gene effect between genes predicted as splice-altered by SpliceAI and the rest"
  )
```
```{r}
ggsave("/Users/castilln/Desktop/thesis/figures/msk_spliceAI_gene_effect.png", width=9)
```
