---
title: "Parse SpliceAI"
output:
  html_document:
    df_print: paged
---

## Load libraries
```{r}
library(tidyverse)
library(readr)
library(data.table)

select = dplyr::select
mutate = dplyr::mutate
rename = dplyr::rename
filter = dplyr::filter
setdiff = dplyr::setdiff
slice = dplyr::slice
```

## Load data

Read data and merge results for each chromosome in a single file
```{r, message=FALSE}
data_files <- list.files("/Users/castilln/Desktop/thesis/localdata/spliceai/output_q4/")

#read all .vcf files
splice_output_list = list()

#obs that chr X has been transformed to chr22 and chr Y has been transformed to chr23
for (i in 1:23) {
  splice_output_list[[i]] <- assign(paste0("splice_chr", i),
      fread(paste0("/Users/castilln/Desktop/thesis/localdata/spliceai/output_q4/", data_files[i]), header = TRUE, sep = "\t"))
}

#bind all outputs together
rbindlist(splice_output_list) -> full_splice_out
full_splice_out %>% 
  rename("CHROM" = "#CHROM") -> full_splice_out

#order chromosomes
ordered_splice <- full_splice_out[order(full_splice_out[,1]),]

```

## Begin parsing

Separate info into columns
```{r}
df_splice =
  ordered_splice %>%
  separate(
    col = INFO,
    sep = "\\|",
    into = c(
      "INFO",
      "SYMBOL",
      "SCORE_ACC_GAIN",
      "SCORE_ACC_LOSS",
      "SCORE_DONOR_GAIN",
      "SCORE_DONOR_LOSS",
      "POS_ACC_GAIN",
      "POS_ACC_LOSS",
      "POS_DONOR_GAIN",
      "POS_DONOR_LOSS",
      "EXTRA"
      ),
    extra = "merge",
    fill = "right"
  )

head(df_splice)
```

Move over all additional splice events into a separate column called ADDITIONAL_EVENTS
```{r}
df_splice =
  df_splice %>% 
  mutate(ADDITIONAL_EVENTS = POS_DONOR_LOSS) %>% 
  mutate(POS_DONOR_LOSS = str_extract(ADDITIONAL_EVENTS,"[^,]*")) %>%
  mutate(ADDITIONAL_EVENTS = str_extract(ADDITIONAL_EVENTS,",.*$")) %>% 
  mutate(ADDITIONAL_EVENTS = substr(ADDITIONAL_EVENTS,2,nchar(ADDITIONAL_EVENTS))) %>% 
  mutate(ADDITIONAL_EVENTS = paste(ADDITIONAL_EVENTS,EXTRA,sep = "|")) %>%
  mutate(ADDITIONAL_EVENTS = gsub("NA\\|NA",NA,ADDITIONAL_EVENTS)) %>% 
  select(-EXTRA)

head(df_splice)
```
Let's store additional splice events as a new tibble
```{r}
df_additional = 
  str_split(df_splice$ADDITIONAL_EVENTS,pattern = ",",simplify = TRUE) %>% 
  as.data.frame() %>% 
  mutate_all(na_if,"")

```

We can now calculate the number of additional splice events and store in our original data
```{r}
no_events = apply(!is.na(df_additional), 1, sum)

df_splice =
  df_splice %>% 
  mutate(NO_ADD_EVENTS = no_events)
```

Let's store additional splice events as a new tibble
```{r}
df_additional = 
  str_split(df_splice$ADDITIONAL_EVENTS,pattern = ",",simplify = TRUE) %>% 
  as.data.frame() %>% 
  mutate_all(na_if,"") %>% 
  filter(!is.na(V1))
```

Let's subset the data by SNPs with multiple events
```{r}
df_splice_add =
  df_splice %>% 
  filter(NO_ADD_EVENTS != 0)
```


Let's create a tibble with the total number of additional events to be added
```{r}
tot_no_events = sum(no_events)

df_add_events = 
  data.frame(matrix(nrow = tot_no_events, ncol = 17))

colnames(df_add_events) = colnames(df_splice)[1:17]
```

Now we can finally loop and add the events to our matrix
```{r}
event_index = 1
for (SNP_index in 1:nrow(df_splice_add)) {
  events_in_SNP = df_splice_add$NO_ADD_EVENTS[SNP_index]
  for (event_index_SNP in 1:events_in_SNP) {
    # Add baseline info for each variant
    df_add_events[event_index,1:7] = df_splice_add[SNP_index,1:7]

    # Extract splice info for each gene variant
    splice_info_str = df_additional[SNP_index,event_index_SNP]
    splice_info_vec = str_split(splice_info_str,pattern = "\\|",simplify = TRUE)
    
    # Add gene specific splice info for each variant
    df_add_events[event_index,8:17] = splice_info_vec
    
    event_index = event_index + 1
  }
}
```

Create result matrix
```{r}
spliceai_out = df_splice %>% 
  as.data.frame()

spliceai_out = spliceai_out[,1:17]

spliceai_out = rbind(spliceai_out,df_add_events)

spliceai_out = 
  spliceai_out %>% 
  arrange(desc(POS))


write.csv(spliceai_out, "~/Desktop/thesis/localdata/spliceai/output/spliceAI_out.csv", quote = FALSE, row.names = FALSE)
```