---
title: "Table off-target SpliceAI events"
output: html_notebook
---

Load libraries 
```{r}
library(tidyverse)
library(data.table)

select = dplyr::select
filter = dplyr::filter
```

Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis")
```

Load linear model & df
```{r}
##DEPENDENCIES
dependencies_df <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_gene_dependency.csv", header = TRUE)

#MUTATIONS AND METADATA
CCLE_mutations <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_info", header = TRUE) 


## READ SPLICEAI RESULTS
predicted_splice <- readRDS("github/results/spliceAI/ths5.rds") 
```


```{r}
head(predicted_splice)
```


From spliceAI predictions, select the off-target mutations
```{r}
predicted_na = 
  predicted_splice %>% 
  filter(off_target ==  1)%>% 
  select(-c("primary_disease", "Variant_Classification", "Annotation_Transcript", "DepMap_ID"))
```

Annotate primary disease
```{r}
head(CCLE_mutations)
lines_dis = 
  CCLE_mutations %>% 
  select(primary_disease, stripped_cell_line_name, DepMap_ID) %>% 
  distinct() %>% 
  dplyr::rename("CELL_LINE_NAME" = "stripped_cell_line_name",
         "DISEASE" = "primary_disease")


predicted_na = 
  predicted_na %>% 
  left_join(lines_dis, by = "CELL_LINE_NAME") %>% 
  as.data.table() %>% 
  distinct()

head(predicted_na)
```

Prepare df to export to table
```{r}
table = 
  predicted_na %>% 
  select(DepMap_ID, SYMBOL, DISEASE) %>% 
  filter(!is.na(DISEASE)) %>% 
  distinct() %>% 
  as_data_frame()
  
head(table)
```

```{r}
library(xtable)

print(xtable(table, floating=FALSE, latex.environments=NULL), include.rownames = F)
```


Create df to plot
```{r}
table_events = 
  predicted_na %>% 
  select(SYMBOL, DISEASE, DepMap_ID) %>% 
  filter(!is.na(DISEASE)) %>% 
  group_by(DISEASE) %>% 
  add_tally(name = "n_events") %>% 
  select(DISEASE, n_events) %>% 
  distinct

head(table_events)
```
```{r}
ggplot(data = table_events, aes(x = reorder(DISEASE, n_events), y = n_events)) + 
  geom_col(fill="black") + 
  coord_flip() + 
  theme_bw() + 
  xlab("") + 
  ylab("Off-target splicing events predicted by SliceAI")

ggsave(filename = "n_off_target_events",plot = last_plot(),path = "figures/results/spliceai/", device = "png", width=5, height=4)
```


Get the name of the cell lines that have off target splice
```{r}
cell_lines_na = 
  predicted_na %>% 
  pull(CELL_LINE_NAME) %>% 
  unique() %>% 
  as.data.frame()

colnames(cell_lines_na) <- "CELL_LINE_NAME"
```

In this way we get a DepMap ID for the SNP, but keep in mind that that DepMap ID refers to other mutation in the same gene. 

Use the cell line ID to extract the DepMap ID
```{r}
na_depmap = 
  CCLE_mutations %>% 
  semi_join(cell_lines_na, by = "stripped_cell_line_name") %>% select(DepMap_ID) %>% 
  unique()
```


Use the DepMap ID to filter those cell lines for which dependencies have been calculated that also had predicted off target splicing
```{r}
filt_dependencies = 
  dependencies_df %>% 
  semi_join(na_depmap, by = "DepMap_ID")
```

Pivot longer & create a new column for entrez ID 
```{r}
long_dependencies = 
  filt_dependencies %>% 
  pivot_longer(cols = -DepMap_ID, names_to = "SYMBOL", values_to = "DEPENDENCY" )

long_dependencies = 
  long_dependencies %>%
    separate(col = "SYMBOL", into = c("SYMBOL", "ENTREZ"), sep = "\\s") 
   
long_dependencies$ENTREZ <- gsub("\\(|\\)","",as.character(long_dependencies$ENTREZ))
```

Find the dependency for the specific gene
```{r}
long_dependencies =
  long_dependencies %>% 
  select(-ENTREZ)

pred_symbol_depmap = 
  predicted_na %>% 
  select(DepMap_ID.y, SYMBOL) %>% 
  distinct() %>% 
  rename("DepMap_ID" = "DepMap_ID.y")

filt_long_dependencies =
  long_dependencies %>% 
  semi_join(pred_symbol_depmap, by = c("DepMap_ID", "SYMBOL")) %>% 
  arrange(desc(DEPENDENCY))

filt_long_dependencies = 
  filt_long_dependencies %>% 
  left_join(lines_dis, by = "DepMap_ID")
```

Save dependency dataframe
```{r}
saveRDS(filt_long_dependencies, file = "../spliceai/Dependencies_offTarget_spliceAI.rds")
```

Plot distribution of the dependencies for these genes 
```{r}
p = ggplot(filt_long_dependencies, aes(x = DISEASE, y = DEPENDENCY)) + 
  geom_boxplot() + 
  coord_flip() + 
  theme_bw() + 
  theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        ) +
  xlab(NULL)
```

Save plot
```{r}
ggsave("../../../figures/results/spliceai/boxplot_offTarget_spliceAI_dependencies.png", height = 7, width = 10)
```