---
title: "SpliceAI analysis - role of transitions and transversions in generation of new splice variants"
output: html_notebook
---

Description: Classification of SNPs based on transitions and transversions and study of whether one of this is more prevalent than the other in the generation of new splicing sites. 

Load libraries 
```{r}
library(tidyverse)
library(data.table)

rename = dplyr::rename
select = dplyr::select
```

set wd 
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```

Load data
```{r}
splice_out_05 = readRDS("spliceai/spliceAI05_Annotated.rds")
splice_out = fread("spliceai/output/spliceAI_out.csv")

head(splice_out)
```

Create new column with highest probability 
```{r}
##TURN SCORE COLUMNS INTO NUMERIC VARIABLES
#SELECT SCORE COLUMNS
score_col = 
  splice_out %>% 
  select(contains("SCORE")) 

#SELECT COLUMN NAMES
cols.num <- colnames(score_col)

#CONVERT INTO NUMERIC
splice_out = as.data.frame(splice_out)
splice_out[cols.num] <- sapply(splice_out[cols.num],as.numeric)

head(splice_out)

##SELECT MAX PROBABILITY FROM SPLICEAI OUTPUT
splice_out_max =
  splice_out%>% 
  mutate(MAX_PROBABILITY = apply(X = splice_out[,10:13], MARGIN = 1, FUN = max)) %>% 
  drop_na(MAX_PROBABILITY)
```

Combine REF and ALT columns 
```{r}
combined_splice = 
  splice_out_max %>% 
  unite(SUBS, c("REF", "ALT"), sep = "-")
  
head(combined_splice)
```

Annotate DNA substitutions as transcriptions or transversions
```{r}
splice_out_substitutions = 
  combined_splice %>% 
  mutate(SUBS = case_when(
    SUBS == "G-A" ~ "G-A", 
    SUBS == "A-G" ~ "G-A", 
    SUBS == "C-T" ~ "C-T",
    SUBS == "T-C" ~ "C-T",
    SUBS == "G-T" ~ "G-T", 
    SUBS == "T-G" ~ "G-T", 
    SUBS == "A-T" ~ "A-T", 
    SUBS == "T-A" ~ "A-T", 
    SUBS == "C-A" ~ "C-A", 
    SUBS == "A-C" ~ "C-A", 
    SUBS == "C-G" ~ "G-C", 
    SUBS == "G-C" ~ "G-C"
    )) %>% 
  drop_na(SUBS)

splice_out_substitutions_ann = 
  splice_out_substitutions %>% 
  mutate(SUBS_ANNOTATION = case_when(
    SUBS == "G-A" ~ "TRANSITION", 
    #SUBS == "A-G" ~ "TRANSITION", 
    SUBS == "C-T" ~ "TRANSITION", 
    #SUBS == "T-C" ~ "TRANSITION",
    SUBS == "G-T" ~ "TRANSVERSION", 
    SUBS == "G-C" ~ "TRANSVERSION",
    SUBS == "A-T" ~ "TRANSVERSION", 
    #SUBS == "A-C" ~ "TRANSVERSION",
    SUBS == "C-A" ~ "TRANSVERSION", 
    #SUBS == "T-A" ~ "TRANSVERSION",
    #SUBS == "C-G" ~ "TRANSVERSION", 
    #SUBS == "T-G" ~ "TRANSVERSION"
  )) %>% 
  mutate(MAX_PROBABILITY = as.numeric(MAX_PROBABILITY)) %>% 
  drop_na(MAX_PROBABILITY)
```

Plot number of each time of substitution 
```{r}
splice_out_substitutions %>% 
  filter(MAX_PROBABILITY > 0.5) %>% 
  ggplot(aes(SUBS)) + 
  geom_bar(fill = "black") + 
  theme_bw() + 
  coord_flip()
```


Plot
```{r}
library(ggpubr)
library(ggbeeswarm)
library(viridis)


splice_out_substitutions_ann %>% 
  filter(MAX_PROBABILITY > 0.5) %>% 
  ggplot(aes(reorder(SUBS, MAX_PROBABILITY), MAX_PROBABILITY, fill=SUBS_ANNOTATION)) +
  geom_boxplot(scale = T) + 
  #geom_jitter(alpha = 0.3, size = 0.3) + 
  theme_bw() +
  scale_fill_viridis_d(labels = c("Transition", "Transversion"), direction = -1) +
  ylab("Probability of generating a splice event") +
  xlab(NULL) + 
  theme(axis.text.y = element_blank(),
        legend.position = "bottom") 
```

` 

