---
title: "Linear model on dependencies - pan cancer and per cancer"
output: html_notebook
---

Description: 

Load libraries
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(readr)
library(data.table)
library(skimr)
library(moderndive)
library(gapminder)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Load data
```{r}
##GENE EFFECT
gene_eff_df <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_gene_effect.csv", header = TRUE) 

##REPLICATE MAP
rep_map <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_replicate_map.csv", header = TRUE)

##DEPENDENCIES
dependencies_df <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_gene_dependency.csv", header = TRUE)

#METADATA
sample_info <- read_csv("/Users/castilln/Desktop/thesis/localdata/depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("/Users/castilln/Desktop/thesis/localdata/depmap/mutations_spliceosome.csv")
```
Create meta data of interest
```{r}
meta = 
  sample_info %>% 
  select(c("DepMap_ID", "stripped_cell_line_name", "primary_disease"))
```


Join tidy crispr data and join with metadata
```{r}
gene_eff_meta = 
  gene_eff_df %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "Gene", values_to = "value") %>% 
  left_join(meta, by = "DepMap_ID")
```

Annotate mutations in the spliceosome
```{r}
gene_eff_mutated = 
 gene_eff_meta %>% 
   mutate(spliceosome_mutated = 
           case_when(
            gene_eff_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !gene_eff_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) 
```


```{r}
table(distinct(gene_eff_mutated)$primary_disease,distinct(gene_eff_mutated)$spliceosome_mutated)
```

## Pre-processing: quality check. 
Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
keep <- c("Bone Cancer", "Bile Duct Cancer", "Leukemia","Pancreatic Cancer")

df_filt = 
  gene_eff_mutated %>% 
  filter(primary_disease %in% keep) %>% 
  drop_na()
```
Create a new column for entrez ID 
```{r}
df_filt = 
  df_filt %>%
    separate(col = "Gene", into = c("Hugo_Symbol", "entrez_id"), sep = "\\s") 
   
df_filt$entrez_id <- gsub("\\(|\\)","",as.character(df_filt$entrez_id))
```

```{r}
head(df_filt)
```

QQ plots
```{r, warning = F}
theme_set(
  theme_minimal() +
    theme(legend.position = "top")
  )

##PLOT 
ggplot(df_filt, aes(sample = value)) + 
  facet_wrap(facets = "primary_disease") +
  stat_qq() + 
  stat_qq_line() 
```

```{r}
df_filt_s = 
  df_filt %>% 
  select(DepMap_ID, Hugo_Symbol, value, spliceosome_mutated, primary_disease) %>% 
  mutate(primary_disease = as.factor(primary_disease),
         spliceosome_mutated = as.factor(spliceosome_mutated))

glimpse(df_filt_s)
```

```{r}
df_filt_s %>% select(value, primary_disease, spliceosome_mutated) %>% skim()
```

```{r, fig.width=5, fig.height=5}
ggplot(df_filt_s, aes(y = value, x = spliceosome_mutated, fill = spliceosome_mutated)) + 
  geom_violin() + 
  geom_jitter(shape=0.1, position=position_jitter(0.2)) + 
  facet_wrap(facets = df_filt_s$primary_disease) + 
  stat_compare_means()
```

```{r, fig.width=5, fig.height=5}
ggplot(df_filt_s, aes(y = value, x = spliceosome_mutated, fill = spliceosome_mutated)) + 
  geom_boxplot(aes(group = spliceosome_mutated), outlier.alpha = 0.7) + 
  facet_wrap(facets = df_filt_s$primary_disease) + 
  stat_compare_means()
```

```{r, fig.width=5, fig.height=5}
library(ggjoy)
ggplot(df_filt_s, aes(x = value, y = spliceosome_mutated)) + 
  geom_joy(scale = 2, alpha=0.5) + 
  theme_joy() +
  facet_grid(facets = df_filt_s$primary_disease)
```


## Wilcoxon-test
Estimate significance using a non-parametric method
Prepare data for Wilcoxon 
```{r, message=FALSE}
df_filt %>% 
  group_by(spliceosome_mutated) %>% 
  summarise(
    count = n(),
    median = median(value, na.rm = TRUE),
    IQR = IQR(value, na.rm = T), 
    sd = sd(value)
    )
```

Visualize data 
```{r, fig.width=2, fig.height=2}
ggplot(df_filt, aes(x = spliceosome_mutated, y = value)) +
  geom_boxplot() + 
  stat_compare_means()
```


2 samples Wilcoxon test
```{r}
res <- wilcox.test(value  ~ spliceosome_mutated, data = df_filt, exact = FALSE)

res
```

Print p-value 
```{r}
res$p.value
```

Wilcoxon test for each gene 
```{r}
library(broom)

wilc_res =
  df_filt %>% 
  ##NEST BY CELL LINE
  nest(-Hugo_Symbol) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
  unnest(tidied) %>% 
  select(-c("data", "model")) 
  
wilc_res
```
FDR - set lower threshold?
```{r}
wilc_res %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(p.value) %>% 
  head()
```

Repeat wilcoxon test per cancer 
```{r}
##LIST OF CANCERS
list_cancer = 
  df_filt %>% 
  pull(primary_disease) %>% 
  unique()

##EMPTY LIST TO STORE LINEAR MODELS
wilc_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  wilc_list[[paste0("wilc.", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-Hugo_Symbol) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```

Pancreatic cancer
```{r}
wilc_list$`wilc.Pancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Leukemia
```{r}
wilc_list$wilc.Leukemia%>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```
Bone cancer
```{r}
wilc_list$`wilc.Bone Cancer`%>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(p.value)
```
Bile duct cancer
```{r}
wilc_list$`wilc.Bile Duct Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```
Repeat the same for t-test
Use Anderson-Darling normality test
```{r}
library(nortest)

ad.test(dis_df$value)
```

```{r}
##EMPTY LIST TO STORE LINEAR MODELS
t_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  t_list[[paste0("t", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-Hugo_Symbol) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~t.test(value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```

Pancreatic cancer
```{r}
t_list$`tPancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Bile duct cancer
```{r}
t_list$`tBile Duct Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```
Bone cancer
```{r}
t_list$`tBone Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```
Leukemia
```{r}
t_list$`tLeukemia` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Try logistic regression? Try to predict spliceosome mutated yes/no from the gene dependencies.