---
title: "Analysis of dependencies"
output: html_notebook
---

Load libraries
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(readr)
library(data.table)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

```{r}
##DEPENDENCIES
dependencies_df <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_gene_dependency.csv", header = TRUE)

#METADATA
sample_info <- read_csv("/Users/castilln/Desktop/thesis/localdata/depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("/Users/castilln/Desktop/thesis/localdata/depmap/mutations_spliceosome.csv")
```

Create meta data of interest
```{r}
meta = 
  sample_info %>% 
  select(c("DepMap_ID", "stripped_cell_line_name", "primary_disease"))
```

Join tidy crispr data and join with metadata
```{r}
dep_meta = 
  dependencies_df %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "GENE", values_to = "DEPENDENCY") %>% 
  left_join(meta, by = "DepMap_ID")
```

Annotate mutations in the spliceosome
```{r}
dep_meta_mut = 
  dep_meta %>% 
   mutate(spliceosome_mutated = 
           case_when(
            dep_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !dep_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) 
```


```{r}
table(distinct(dep_meta_mut)$primary_disease,distinct(dep_meta_mut)$spliceosome_mutated)
```

## Pre-processing: quality check. 
Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
keep <- c("Bone Cancer", "Bile Duct Cancer", "Leukemia","Pancreatic Cancer", "Ovarian Cancer")

df_filt = 
  dep_meta_mut %>% 
  filter(primary_disease %in% keep) %>% 
  drop_na()
```

Create a new column for entrez ID 
```{r}
df_filt = 
  df_filt %>%
    separate(col = "GENE", into = c("HUGO_SYMBOL", "ENTREZ"), sep = "\\s") 
   
df_filt$ENTREZ <- gsub("\\(|\\)","",as.character(df_filt$ENTREZ))
```

QQ plots
```{r, warning = F}
theme_set(
  theme_minimal() +
    theme(legend.position = "top")
  )

##PLOT 
ggplot(df_filt, aes(sample = DEPENDENCY)) + 
  facet_wrap(facets = "primary_disease") +
  stat_qq() + 
  stat_qq_line() 
```
## Wilcoxon-test
Estimate significance using a non-parametric method
Prepare data for Wilcoxon 
```{r, message=FALSE}
df_filt %>% 
  group_by(spliceosome_mutated) %>% 
  summarise(
    count = n(),
    median = median(DEPENDENCY, na.rm = TRUE),
    IQR = IQR(DEPENDENCY, na.rm = T),
    std = sd(DEPENDENCY, na.rm = T)
    )
```

Visualize data 
```{r}
ggboxplot(df_filt, x = "spliceosome_mutated", y = "DEPENDENCY", 
          color = "spliceosome_mutated", palette = c("mediumvioletred","midnightblue"), 
          ylab = "Dependency", xlab = "Defective spliceosome")
```

2 samples Wilcoxon test
```{r}
res <- wilcox.test(DEPENDENCY  ~ spliceosome_mutated, data = df_filt, exact = FALSE)

res
```

Nest and perform wilcooxn test per gene
```{r}
library(broom)
wilc_res =
  df_filt %>% 
  ##NEST BY CELL LINE
  nest(-HUGO_SYMBOL) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(DEPENDENCY ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
  unnest(tidied) %>% 
  select(-c("model", "data"))
  
wilc_res
```

FDR 
```{r}
wilc_res %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr) %>% 
  head()
```

Repeat wilcoxon test per cancer 
```{r}
##LIST OF CANCERS
list_cancer = 
  df_filt %>% 
  pull(primary_disease) %>% 
  unique()

##EMPTY LIST TO STORE LINEAR MODELS
lm_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  lm_list[[paste0("lm.", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-HUGO_SYMBOL) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(DEPENDENCY ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```

Pancreatic cancer
```{r}
lm_list$`lm.Pancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Ovarian cancer
```{r}
lm_list$`lm.Ovarian Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Bone cancer
```{r}
lm_list$`lm.Bone Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Bile duct cancer
```{r}
lm_list$`lm.Bile Duct Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Leukemia
```{r}
lm_list$lm.Leukemia %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Repeat wilcoxon test per cancer 
```{r}
library(mblm)
##LIST OF CANCERS
list_cancer = 
  df_filt %>% 
  pull(primary_disease) %>% 
  unique()

df_filt_mb = 
  df_filt %>% 
  mutate(spliceosome = case_when(
    spliceosome_mutated == "YES" ~ 1, 
    spliceosome_mutated == "NO" ~ 0
  ))

##EMPTY LIST TO STORE LINEAR MODELS
mb_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt_mb %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  mb_list[[paste0("mb.", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-HUGO_SYMBOL) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~mblm(DEPENDENCY ~ spliceosome, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```