---
title: "Gene effect - pan cancer and per cancer - transformed data"
output: html_notebook
---

Load libraries
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(readr)
library(data.table)
library(skimr)
library(moderndive)
library(gapminder)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Load data
```{r}
##GENE EFFECT
gene_eff_df <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_gene_effect.csv", header = TRUE) 

#METADATA
sample_info <- read_csv("/Users/castilln/Desktop/thesis/localdata/depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("/Users/castilln/Desktop/thesis/localdata/depmap/mutations_spliceosome.csv")
```

Create meta data of interest
```{r}
meta = 
  sample_info %>% 
  select(c("DepMap_ID", "stripped_cell_line_name", "primary_disease"))
```


Join tidy crispr data and join with metadata
```{r}
gene_eff_meta = 
  gene_eff_df %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "Gene", values_to = "value") %>% 
  left_join(meta, by = "DepMap_ID")
```

Annotate mutations in the spliceosome
```{r}
gene_eff_mutated = 
 gene_eff_meta %>% 
   mutate(spliceosome_mutated = 
           case_when(
            gene_eff_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !gene_eff_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) 
```


```{r}
table(distinct(gene_eff_mutated)$primary_disease,distinct(gene_eff_mutated)$spliceosome_mutated)
```

## Pre-processing: quality check. 
Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
keep <- c("Bone Cancer", "Bile Duct Cancer", "Leukemia","Pancreatic Cancer")

df_filt = 
  gene_eff_mutated %>% 
  filter(primary_disease %in% keep) %>% 
  drop_na()
```

Create a new column for entrez ID 
```{r}
df_filt = 
  df_filt %>%
    separate(col = "Gene", into = c("Hugo_Symbol", "entrez_id"), sep = "\\s") 
   
df_filt$entrez_id <- gsub("\\(|\\)","",as.character(df_filt$entrez_id))
```

```{r}
head(df_filt)
```

## Square root transformation
```{r}
norm_df_filt = 
  df_filt %>% 
  mutate(trans_value = (sign(value) * abs(value)^(1/3))) %>% 
  mutate(primary_disease = as.factor(primary_disease))

head(norm_df_filt)
```

```{r}
ggplot(norm_df_filt, aes(sample = trans_value)) + 
  facet_wrap(facets = "primary_disease") +
  stat_qq() + 
  stat_qq_line() 
```

```{r}
ggplot(norm_df_filt, aes(y = trans_value, x = spliceosome_mutated, fill = spliceosome_mutated)) + 
  geom_boxplot() + 
  facet_wrap(facets = norm_df_filt$primary_disease)
```

```{r}
norm_df_filt %>% skim()
```

```{r}
library(ggjoy)
ggplot(norm_df_filt, aes(x = trans_value, y = spliceosome_mutated)) + 
  geom_joy(scale = 2, alpha=0.5) + 
  theme_joy() +
  facet_grid(facets = norm_df_filt$primary_disease)
```

Use SAnderson-Darling normality test
```{r}
library(nortest)

ad.test(norm_df_filt$trans_value)
```
```{r}
head(norm_df_filt)
```


Repeat wilcoxon test per cancer 
```{r}
library(broom)
##LIST OF CANCERS
list_cancer = 
  norm_df_filt %>% 
  pull(primary_disease) %>% 
  unique()

##EMPTY LIST TO STORE LINEAR MODELS
wilc_list <- list()

for (dis in list_cancer){
  dis_df = 
    norm_df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  wilc_list[[paste0("wilc.", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-Hugo_Symbol) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~wilcox.test(trans_value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```

Pancreatic cancer
```{r}
wilc_list$`wilc.Pancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Ovarian cancer
```{r}
wilc_list$`wilc.Bone Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```

Bile Duct cancer
```{r}
wilc_list$`wilc.Bile Duct Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(p.value)
```


```{r}
wilc_list$`wilc.Leukemia` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```


Repeat analysis per cancer - t-test
```{r}
##EMPTY LIST TO STORE LINEAR MODELS
t_list <- list()

for (dis in list_cancer){
  dis_df = 
    norm_df_filt %>% 
    filter(primary_disease == dis)
  
  ##LINEAR MODEL
  t_list[[paste0("t.", dis)]] = 
  dis_df %>% 
  ##NEST BY CELL LINE
  nest(-Hugo_Symbol) %>% 
  #LM - DESIGN
  mutate(model = map(data, ~t.test(trans_value ~ spliceosome_mutated, data = .))) %>% 
  #TIDY THE MODEL
  mutate(tidied = map(model, tidy)) %>% 
    unnest(tidied) %>% 
    select(-c("data", "model"))
}
```
Pancreatic cancer
```{r}
t_list$`t.Pancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p.value, method = "fdr")) %>% 
  arrange(fdr)
```
