---
title: "Kolmogorov-Smirnov test in data from CRISPr screening"
output: html_notebook
---

Description: test statistical significance of dependencies between spliceosome-wt cell lines and spliceosome-mutated cell lines using K-S test. 

Libraries 
```{r}
library(tidyverse)
library(edgeR)
library(ggpubr)
library(readr)
library(data.table)
library(moderndive)
library(gapminder)
library(dgof)
select = dplyr::select
rename = dplyr::rename
filter = dplyr::filter
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Data
```{r, message=FALSE}
##GENE EFFECT
gene_eff_df <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/achilles/Achilles_gene_effect.csv", header = TRUE) 

#METADATA
sample_info <- read_csv("/Users/castilln/Desktop/thesis/localdata/depmap/sample_info.csv") #metadata

#MUTATIONS AND SAMPLE INFO
CCLE_mutations <- fread("/Users/castilln/Desktop/thesis/localdata/depmap/CCLE_info", header = TRUE) 

#LIST OF MUTATIONS FROM CCLE MUTATIONS THAT AFFECT THE SPLICEOSOME
mutations_spliceosome = fread("/Users/castilln/Desktop/thesis/localdata/depmap/mutations_spliceosome.csv")
```

Create meta data of interest
```{r}
meta = 
  sample_info %>% 
  select(c("DepMap_ID", "stripped_cell_line_name", "primary_disease"))
```


Join tidy crispr data and join with metadata
```{r}
gene_eff_meta = 
  gene_eff_df %>% 
  pivot_longer(cols=-DepMap_ID, names_to = "Gene", values_to = "value") %>% 
  left_join(meta, by = "DepMap_ID")

head(gene_eff_meta)
```

Annotate mutations in the spliceosome
```{r}
gene_eff_mutated = 
 gene_eff_meta %>% 
   mutate(spliceosome_mutated = 
           case_when(
            gene_eff_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "YES", #if the mutation is in a gene from the spliceosome: 1 
            !gene_eff_meta$stripped_cell_line_name %in% mutations_spliceosome$stripped_cell_line_name ~ "NO")) 

head(gene_eff_mutated)
```
```{r}
table(distinct(gene_eff_mutated)$primary_disease,distinct(gene_eff_mutated)$spliceosome_mutated)
```

## Pre-processing
Let's keep only those cancers with a relatively significant number of cell lines w/o mutations in the spliceosome:
```{r}
keep <- c("Bone Cancer", "Bile Duct Cancer", "Leukemia","Pancreatic Cancer", "Leukemia")

df_filt = 
  gene_eff_mutated %>% 
  filter(primary_disease %in% keep) %>% 
  drop_na()
```

Create a new column for entrez ID 
```{r}
df_filt = 
  df_filt %>%
    separate(col = "Gene", into = c("Hugo_Symbol", "entrez_id"), sep = "\\s") 
   
df_filt$entrez_id <- gsub("\\(|\\)","",as.character(df_filt$entrez_id))
```

```{r}
head(df_filt)
```
Transform primary disease as factor
```{r}
df_filt = 
  df_filt %>% 
  mutate(primary_disease = as.factor(primary_disease))
```

Let's draw the distribution for each cancer
```{r, fig.width=5, fig.height=5}
ggplot(df_filt, aes(y = value, x = spliceosome_mutated, fill = spliceosome_mutated)) + 
  geom_violin() + 
  geom_jitter(size=0.01, position=position_jitter(0.2), alpha = 0.4) +
  scale_fill_manual(values = c("NO" = "gray", "YES" = "red")) +
  facet_wrap(facets = df_filt$primary_disease) + 
  stat_compare_means() + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        )
```

Kolmogorov-Smirnov test
```{r}
#SEPARATE THE DATAFRAME INTO TWO GROUPS
df_no = df_filt[df_filt$spliceosome_mutated == "NO",]
df_yes = df_filt[df_filt$spliceosome_mutated == "YES",]

#WT SPLICEOSOME
#TRASNFORM TO WIDE FORMAT (SYMBOLS AS COLUMNS)
df_no_wide = 
  df_no %>% 
  pivot_wider(id_cols = DepMap_ID, names_from = Hugo_Symbol) %>% 
  as.data.frame()

#DROP NA VALUES
df_no_wide <- drop_na(df_no_wide)

#SET DEPMAP ID AS ROWNAMES
rownames(df_no_wide) <- df_no_wide$DepMap_ID
df_no_wide <- df_no_wide[,-1]

#MUTATED SPLICEOSOME
#TRASNFORM TO WIDE FORMAT (SYMBOLS AS COLUMNS)
df_yes_wide = 
  df_yes %>% 
  pivot_wider(id_cols = DepMap_ID, names_from = Hugo_Symbol) %>% 
  as.data.frame()

#DROP NA VALUES
df_yes_wide <- drop_na(df_yes_wide)

#SET DEPMAP ID AS ROWNAMES
rownames(df_yes_wide) <- df_yes_wide$DepMap_ID
df_yes_wide <- df_yes_wide[,-1]


df_ks <- data.frame(Symbol = character(18119), D = numeric(18119), p = numeric(18119), stringsAsFactors = F)
```

Loop for KS test --- pan-cancer
```{r}
for (j in 1:18119){  
  k <- ks.test(df_no_wide[,j], df_yes_wide[,j], exact = T)  
  df_ks$Symbol[j] <- names(df_no_wide)[j]
  df_ks$D[j] <- k$statistic
  df_ks$p[j] <- k$p.value
}
```

```{r}
head(df_ks)
```

FDR correction
```{r}
df_ks_fdr = 
  df_ks %>%
  mutate(FDR=p.adjust(p, method = "fdr")) %>% 
  filter(FDR < 0.05)

head(df_ks_arranged)
```

Plot 30 genes with smallest FDR value
```{r}
df_ks_fdr %>% 
  slice_min(order_by = FDR,  n = 10)
```



KS test per cancer 
```{r}
head(df_filt)
```

```{r}
##LIST OF CANCERS
list_cancer = 
  df_filt %>% 
  pull(primary_disease) %>% 
  unique()

##EMPTY LIST TO STORE LINEAR MODELS
lm_list <- list()

for (dis in list_cancer){
  dis_df = 
    df_filt %>% 
    filter(primary_disease == dis)
  
  #PREPARE OUTPUT DF
  data <- data.frame(Symbol = character(18119), D = numeric(18119), p = numeric(18119), stringsAsFactors = F)

  #SEPARATE THE DATAFRAME INTO TWO GROUPS
  df_no = dis_df[dis_df$spliceosome_mutated == "NO",]
  df_yes = dis_df[dis_df$spliceosome_mutated == "YES",]

  #WT SPLICEOSOME
  #TRASNFORM TO WIDE FORMAT (SYMBOLS AS COLUMNS)
  df_no_wide = 
    df_no %>% 
    pivot_wider(id_cols = DepMap_ID, names_from = Hugo_Symbol) %>% 
    as.data.frame()

  #DROP NA VALUES
  df_no_wide <- drop_na(df_no_wide)

  #SET DEPMAP ID AS ROWNAMES
  rownames(df_no_wide) <- df_no_wide$DepMap_ID
  df_no_wide <- df_no_wide[,-1]

  #MUTATED SPLICEOSOME
  #TRASNFORM TO WIDE FORMAT (SYMBOLS AS COLUMNS)
  df_yes_wide = 
    df_yes %>% 
    pivot_wider(id_cols = DepMap_ID, names_from = Hugo_Symbol) %>% 
    as.data.frame()

  #DROP NA VALUES
  df_yes_wide <- drop_na(df_yes_wide)

  #SET DEPMAP ID AS ROWNAMES
  rownames(df_yes_wide) <- df_yes_wide$DepMap_ID
  df_yes_wide <- df_yes_wide[,-1]

  #KS TEST
  for (j in 1:18119){  
  k <- ks.test(df_no_wide[,j], df_yes_wide[,j],exact = T)  
  data$Symbol[j] <- names(df_no_wide)[j]
  data$D[j] <- k$statistic
  data$p[j] <- k$p.value
  
  }
  
  #CHANGE NAME OF THE DATAFRAME TO THE DISEASE THAT THE LOOP IS CONSIDERING IN EACH ITERATION
  assign(paste("df_ks",dis, sep = "_"), data)
}
```

##Multiple testing correction
Bone cancer 
```{r}
`df_ks_Bone Cancer` %>%
  mutate(fdr=p.adjust(p, method = "fdr")) %>% 
  filter(fdr < 0.05) %>% 
  arrange(fdr)
```
Pancreatic cancer
```{r}
`df_ks_Pancreatic Cancer` %>% 
  mutate(fdr=p.adjust(p, method = "fdr")) %>% 
  filter(fdr < 0.05)
```
OBS. we are looking for differences in the CRISPR data, not in the expression data. Therefore differences mean that the effect of that particular gene in the cell line differed significantly based on whether or not the spliceosome presented mutations. 

Leukemia 
```{r}
df_ks_Leukemia %>% 
  mutate(fdr=p.adjust(p, method = "fdr")) %>% 
  filter(fdr < 0.05) %>% 
  #SINCE ALL GENES SHOW THE SAME P-VALUE, GET RANDOM 30 GENES TO PLOT 
  slice_sample(n = 30) 

to_plot = 
  df_ks_Leukemia %>% 
  mutate(fdr=p.adjust(p, method = "fdr")) %>% 
  filter(fdr < 0.05) %>% 
  #SINCE ALL GENES SHOW THE SAME P-VALUE, GET RANDOM 30 GENES TO PLOT 
  arrange(fdr) %>% 
  slice_head(n = 30)

df_ks_Leukemia_fdr =
  df_ks_Leukemia %>% 
  mutate(fdr=p.adjust(p, method = "fdr"))

df_ks_Leukemia_fdr %>% 
  mutate(is_sig = fdr<0.05) %>% 
  summarise(sig = sum(is_sig)/length(is_sig))

df_ks %>% 
  mutate(fdr=p.adjust(p, method = "fdr")) %>% 
  mutate(is_sig = fdr<0.05) %>%
  summarise(sig = sum(is_sig)/length(is_sig))


head(to_plot)
dim(to_plot)
```

Plot genes with significant differences
```{r, fig.height=8, fig.width=9}                                    
symbols_to_plot = 
  to_plot %>% 
  select(Symbol) %>% 
  as_vector()

df_leuk_plot = 
  df_leuk %>% 
  #PLOT ONLY 30 GENES WITH THE SMALLEST P-VALUE
  filter(df_leuk$Hugo_Symbol %in% symbols_to_plot) 
  
#change font size of wilcoxon test in plots
font_size <- 2
  
#  df_leuk_plot %>%
    
to_plot_pan =     
  df_ks %>% 
  mutate(fdr=p.adjust(p, method = "fdr")) %>% 
  mutate(is_sig = fdr<0.05) %>% 
  arrange(fdr) %>%  
  top_n(n = 30) %>% 
  select(Symbol) %>% 
  as_vector() 

df_pan_plot = 
  df_filt %>% 
  filter(df_filt$Hugo_Symbol %in% to_plot_pan)

df_pan_plot %>% 
  ggplot(aes(y = value, x = spliceosome_mutated, fill = spliceosome_mutated)) + 
  geom_violin() + 
  geom_jitter(size = 0.001, position = position_jitter(0.2), alpha = 0.4) + 
  scale_fill_manual(values = c("NO" = "gray", "YES" = "red")) +
  facet_wrap(~Hugo_Symbol) + 
  stat_compare_means(label.y = 0.8, size = font_size) + 
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        plot.title = element_text(face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        )
```


Bile Duct Cancer
```{r}
`df_ks_Bile Duct Cancer` %>% 
  mutate(fdr=p.adjust(p, method = "fdr")) %>% 
  filter(fdr < 0.05)
```

