---
title: "Gene effect: filtered out IN/DELs, no MSK"
output: html_notebook
---

Libraries
```{r}
library(tidyverse)
library(data.table)
library(ggbeeswarm)
library(readr)
library(ggpubr)


select = dplyr::select
rename = dplyr::rename
```

source R profile. Memory was set to 500000.
```{r}
Sys.setenv("R_ENVIRON_USER"='/Users/castilln/.Renviron')
Sys.getenv("R_ENVIRON_USER")

```

Set wd 
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/castilln/Desktop/thesis/localdata")
```

Load data
```{r}
#DEPMAP somatic mutations
ccle = fread("depmap/ccle_info_21q1.csv")
ccle = 
  ccle %>% 
  rename("SYMBOL" = "Hugo_Symbol")

#CRISPR gene effect
gene_effect = fread("depmap/achilles/Achilles_gene_effect.csv")

#LOAD ACTIONABLE SPLICE OUTPUT (ALREADY FILTERED FOR MSK) 
splice_genes = readRDS("spliceai/spliceAI05_Annotated.rds")
head(splice_genes)
```


Extract MSK IMPACT genes present in the spliceAI output
(This data frame has already been filtered for MSK IMPACT genes)
```{r}
list_spliced_genes =
  splice_genes %>% 
  select(SYMBOL) %>% 
  distinct() %>% 
  pull()

head(list_spliced_genes)
```


Select variables of interest from ccle and filter out MSK IMPACT genes
```{r}
df_mutation_meta = 
  ccle %>% 
  select(DepMap_ID, SYMBOL, primary_disease, Variant_Type, isDeleterious) %>% 
  filter(SYMBOL %in% list_spliced_genes)
head(df_mutation_meta)
```

Create new column with custom isDeleterious information. isDeleterious only if mutation is an insertion or deletion
```{r}
df_mutation_delet = 
  df_mutation_meta %>% 
  mutate(isDel = ifelse(Variant_Type == "SNP", FALSE, TRUE)) %>% 
  select(-isDeleterious)

dim(df_mutation_delet)
head(df_mutation_delet)
```

Pivot long effect matrix and filter for MSK IMPACT genes
```{r}
##PIVOT LONG EFFECT MATRIX
df_crispr = 
  gene_effect %>% 
  pivot_longer(cols = -DepMap_ID, names_to = "SYMBOL", values_to = "gene_effect") %>% 
  mutate(SYMBOL = gsub(" .*","",SYMBOL)) %>% 
  filter(SYMBOL %in% list_spliced_genes)

head(df_crispr)
dim(df_crispr)
```

Add splice variant annotation to spliceAI output data frame
Since this data has already been filtered for SNPs producing a splice variant with >0.5 confidence, as predicted by the NN, we can simply tag all SNPs as TRUE
```{r, fig.height=10, fig.width=10}
df_splice = 
  splice_genes %>% 
  mutate(splice_variant = TRUE,
         splice_gene = SYMBOL) %>%
  select(DepMap_ID,SYMBOL,primary_disease,splice_variant,splice_gene,MAX_PROBABILITY, exon_intron)

head(df_splice)
dim(df_splice)
```

Filter out SNPs without cell line annotation (this is due to different versions of CCLE data being used)
```{r}
df_splice = 
  df_splice %>% 
  filter(!is.na(DepMap_ID))
dim(df_splice)
```

Create new column with custom isDeleterious information. isDeleterious only if mutation is an insertion or deletion
```{r}
df_mutation_delet = 
  df_mutation_meta %>% 
  mutate(isDel = ifelse(Variant_Type == "SNP", FALSE, TRUE)) %>% 
  select(-isDeleterious)

dim(df_mutation_delet)
head(df_mutation_delet)
```



Now, let's join the spliceAI output with the CCLE mutation annotation information.
```{r}
df_splice_annot = 
  df_splice %>% 
  left_join(df_mutation_delet, by = c("SYMBOL", "DepMap_ID", "primary_disease")) %>% 
  distinct() %>% 
  #indicate those splice variants that occur in a different gene than the one we gave as input
  mutate(off_target = ifelse(is.na(Variant_Type), TRUE, FALSE)) %>% 
  #Substitute NAs from Variant_Type with "SNP"
  mutate(Variant_Type = ifelse(is.na(Variant_Type), "SNP", Variant_Type)) 

df_off_target_nomsk = 
  df_splice_annot %>% 
  filter(off_target == T) %>% 
  select(-c("Variant_Type", "isDel", "splice_gene"))

view(df_off_target_nomsk)
#write_csv(df_off_target_nomsk, "spliceai/off_targe_spliceAIout.csv")

dim(df_splice_annot)
head(df_splice_annot)
```


Let's process the mutation metadata in order to tag each gene in each cell line if at least one of the variants in that gene is deleterious.
```{r}
df_mutation_delet_tag =
  df_mutation_delet %>% 
  group_by(DepMap_ID, SYMBOL) %>% 
  summarise(gene_deleterious = ifelse(sum(isDel) > 0,TRUE,FALSE)) %>% 
  ungroup()

head(df_mutation_delet_tag)
dim(df_mutation_delet_tag)
```

Now we can join the deleterious mutation annotated genes and cell lines with the spliceAI output
```{r}
df_splice_deleterious = 
 df_splice_annot %>% 
  left_join(df_mutation_delet_tag, by = c("SYMBOL", "DepMap_ID")) %>% 
  distinct()

head(df_splice_deleterious)
dim(df_splice_deleterious)

df_splice_deleterious_off = 
  df_splice_deleterious %>% 
  filter(off_target == T)

dim(df_splice_deleterious_off)
```
Add disease info to crispr data
```{r}
disease_ccle = 
  ccle %>% 
  select(primary_disease, DepMap_ID) %>% 
  distinct()
```

Finally, we can join this with the CRISPR knockdown data
```{r, fig.height=10, fig.width=10}
df_crispr_splice_annot = 
  df_splice_deleterious_off %>% 
  full_join(df_crispr, by = c("DepMap_ID", "SYMBOL")) %>% 
  # Cell lines without mutations in the MSK IMPACT genes are annotated as FALSE 
  # in both of the splice_variant and gene_deleterious columns
  mutate(splice_variant = ifelse(is.na(splice_variant),FALSE,splice_variant)) %>% 
  mutate(gene_deleterious = ifelse(is.na(gene_deleterious),FALSE, gene_deleterious)) %>%
  # Filter out cell lines without CRISPR knockdown effect data
  filter(!is.na(gene_effect)) %>% 
  mutate(isDel = ifelse(is.na(isDel), FALSE, isDel))


#df_crispr_splice_annot %>% 
 # filter(is.na(primary_disease))

#Add disease information for the SNPs coming from gene effect dataframe 
df_crispr_splice_annot = 
  df_crispr_splice_annot %>% 
  left_join(disease_ccle, by = "DepMap_ID") %>% 
  select(-primary_disease.x) %>% 
  rename("primary_disease" = "primary_disease.y") %>% 
  mutate(variant = case_when(
    splice_variant == TRUE ~ "Var", 
    splice_variant == FALSE ~ "WT"
  ))

sum(df_crispr_splice_annot$variant == "Var")
```

```{r, fig.height=10, fig.width=10}
df_crispr_splice_annot %>% 
  group_by(SYMBOL) %>% 
  mutate(sum_splice_variants = sum(splice_variant)) %>% 
  ungroup() %>%
  filter(sum_splice_variants >= 5) %>% 
  ggplot(aes(y = gene_effect, x = variant, color = variant)) + 
  geom_quasirandom(method = "pseudorandom",alpha = 0.5, size = 0.5) + 
  theme_bw() +
  facet_wrap("SYMBOL") +
  scale_color_manual(values = c("red","black")) +
  xlab("") +
  ylab("CRISPR gene effect (sd from median)") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") + 
  stat_compare_means(label = "p.format", label.y = 4.5, label.x = 1)

#ggsave(filename = "../figures/results/msk_impact/gene_effect/spliceAI_out_05_actionable_geneEffect_filteroutINDELs_wilcoxon.png", height = 10, width = 10)
```

```{r, fig.height=10, fig.width=10}
df_crispr_splice_annot %>% 
  filter(gene_deleterious == FALSE) %>%
  group_by(SYMBOL) %>% 
  mutate(sum_splice_variants = sum(splice_variant)) %>% 
  ungroup() %>%
  filter(sum_splice_variants > 3) %>% 
  ggplot(aes(y = gene_effect, x = variant, color = variant)) + 
  geom_quasirandom(alpha = 0.5, size = 0.5) + 
  theme_bw() +
  facet_wrap("primary_disease") +
  scale_color_manual(values = c("red","black")) +
  xlab("") +
  ylab("CRISPR gene effect (sd from median)") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") + 
  stat_compare_means(label = "p.format", label.y = 4.5, label.x = 1)
```
